---
title: "Transportation Emissions in New England"
author: "Marcos Luna"
date: "1/22/2020"
output: 
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: yes        
    includes:  
      in_header: my_header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of transportation-related emissions in New England

This is an analysis of transportation-related emissions and related externalities in New England.

## Analysis of PM~2.5~ in New England

This is an analysis of PM~2.5~ across New England and its relationship to populations of concern. The analysis is based on data from the EPA's EJSCREEN. EJSCREEN data provides PM~2.5~ annual concentrations at the Census Block Group level for the years 2011 tp 2016 (as of December 2019). 

PM~2.5~ levels vary significantly across New England, with highest concentrations in southwest Connecticut, from Stamford to Hartford, and also from Providence, Rhode Island, to the Boston metro region, and from Portland, Maine along the I-95 corridor to Bangor (see Figure \@ref(fig:mapPM25NE) below). 

```{r mapPM25NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of 2016 annual PM2.5 concentrations across New England at Census Block Group level.", strip.white=TRUE}
library(tidyverse)
library(sf)
library(tmap)
library(tmaptools)
library(maptools)
# library(raster)
library(rgdal)
library(RColorBrewer)
library(sp)
library(CGPfunctions) # for slope graphs
library(ggcorrplot)
library(spdep)
# ne_blkgrp_sf %>% 
#   as.data.frame() %>% 
#   transmute(`PM2.5 2011` = pm_15, `PM2.5 2016` = PM25_19) %>% 
#   summary()

# Load data
load("DATA/ne_layers.rds")

# Map of PM2.5 across New England
tm_shape(ne_blkgrp_sf, unit = "mi") + 
  tm_fill("PM25_19", style = "quantile", 
          title = expression(paste
                             (PM[2.5]," (", mu, "g/", m^3, ")", sep = "")),
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=2)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Annual PM2.5 \nConcentrations\n2016", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

PM~2.5~ concentrations exhibit spatial clustering of both hot spots (i.e. geographic clusters of high values) and cold spots (i.e. geographic clusters of lower values). The map below (Figure \@ref(fig:hotspotPM25NE)) shows statistically significant PM~2.5~ hot spots.

```{r hotspotPM25NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Hot spot map of 2016 annual PM2.5 concentrations at Census Block Group level."}
# Get rid of empty geometries and NAs, and convert to spdf
empty_geo <- st_is_empty(ne_blkgrp_sf)
ne_blkgrp_sp_DemoEJ <- ne_blkgrp_sf[!empty_geo,] %>% 
  dplyr::select(GEOID,PM25_19) %>% 
  # st_transform(., crs = 2163) %>% # convert to US National Atlas Equal Area
  na.omit() %>% 
  as_Spatial()
# Calculate Queen's case neighbors
neighborsQC <- poly2nb(ne_blkgrp_sp_DemoEJ, queen = TRUE)
# Compute neighbor weights
# spdep::set.ZeroPolicyOption(TRUE)
listw <- nb2listw(neighborsQC, style = "W", zero.policy = TRUE)
# compute Getis-Ord Gi statistic
local_g <- localG(ne_blkgrp_sp_DemoEJ$PM25_19, listw)
local_g <- cbind(ne_blkgrp_sp_DemoEJ, as.matrix(local_g))
names(local_g)[3] <- "gstat"
# map the results
tm_shape(local_g, unit = "mi",) + 
  tm_fill("gstat",
          palette = "-RdBu", 
          style = "pretty",
          title = expression(paste("Getis-Ord ", G[i]^"*")),
          showNA = FALSE,
          midpoint = NA,
          labels = c("Significant Clusters","of Low Values","Significant Clusters","of High Values")) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Hot Spot Map of \nPM2.5 for\nNew England", frame = FALSE, main.title.size = 0.6,
            legend.position = c(.8,.2),
            legend.title.size = 0.7)
```

There is a statistically significant hot spot of PM~2.5~ around Stamford, Connecticut. Warm clusters encompass southwestern  Connecticut to Hartford, from Providence, Rhode Island ot Boston, and Augusta to Bangor, Maine along the I-95 corridor. 

These PM~2.5~ levels vary significantly by state (see Figure \@ref(fig:boxplotPM25NE) below). Connecticut stands out with the highest median PM~2.5~ concentrations among the New England states, while Vermont has the lowest overall concentrations. The differences in concentrations between the states are all statistically significant.  

```{r boxplotPM25NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Boxplot of 2016 annual PM2.5 concentrations by state at Census Block Group level."}
# boxplot of PM2.5 by state for 2016
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  ggplot(aes(x = STATE, y = PM25_19, fill = STATE)) + 
  geom_boxplot(notch = TRUE) + 
  ggtitle(expression(atop(paste(PM[2.5], " Annual Concentration by New England state,"), "2016"))) +
  theme_minimal() +
  theme(legend.position = "none", axis.text=element_text(size=8)) + 
  xlab(NULL) + 
  ylab(expression(paste(PM[2.5]," (", mu, "g/", m^3, ")", sep = ""))) +
  coord_flip()
# # boxplot of PM2.5 by state for 2011
# ne_blkgrp_sf %>% 
#   as.data.frame() %>% 
#   ggplot(aes(x = STATE, y = pm_15, fill = STATE)) + 
#   geom_boxplot(notch = TRUE) + 
#   ggtitle(expression(paste(PM[2.5], " Annual Concentration by New England state, 2011", sep = ""))) +
#   theme_minimal() +
#   theme(legend.position = "none") + xlab(NULL) + 
#   ylab(expression(paste(PM[2.5]," (", mu, "g/", m^3, ")", sep = "")))
```


Since 2011, PM~2.5~ levels have declined across the region, on average by `r {round(mean(ne_blkgrp_sf$PM25_pctChange, na.rm = TRUE),1)}`%. Unsurprisingly, this decline has not been uniform (see Figure \@ref(fig:mapPM25NEchange) below). The greatest declines, up to 43%, have been in the northern half of Vermont and the region between Boston and Portsmouth. 

```{r mapPM25NEchange, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of percent change in annual PM2.5 concentrations across New England between 2011 and 2016 at Census Block Group level."}
# Map of PM2.5 change across New England
tm_shape(ne_blkgrp_sf, unit = "mi") + 
  tm_fill("PM25_pctChange", palette = "Greens",
          style = "quantile", 
          title = "% Change",
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=1)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Change in annual PM2.5 \nConcentrations \n2011-2016", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

Averaged across the states, the differences in these declines are also apparent. Figure \@ref(fig:slopegraphPM25NE) below compares the average annual PM~2.5~ concentrations by state and for the region between 2011 and 2016. The region and the states all showed significant declines since 2011, although note that concentrations in Connecticut have remained consistently well above the rest of the region. 

```{r slopegraphPM25NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Slope graph of annual PM2.5 concentrations for 2011 and 2016 for New England and states."}
# Slope graph of PM2.5 by state and region between 2011 and 2016
# Create df of 2019 actual values
PM19wSTAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(PM25Mean = mean(PM25_19, na.rm = TRUE),
            PM25wMean = weighted.mean(x = PM25_19,
                                      w = totalpopE, na.rm = TRUE)) %>% 
  mutate(Year = 2016)
# Create df of 2015 values
PM15wSTAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(PM25Mean = mean(pm_15, na.rm = TRUE),
            PM25wMean = weighted.mean(x = pm_15,
                                      w = pop_15, na.rm = TRUE)) %>% 
  mutate(Year = 2011)
# Create regional average benchmark
pm15NEavg <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  summarize(PM25Mean = mean(pm_15, na.rm = TRUE),
            PM25wMean = weighted.mean(x = pm_15,
                                      w = pop_15, na.rm = TRUE)) %>% 
  transmute(STATE = "NEW ENGLAND", 
            PM25Mean = PM25Mean, 
            PM25wMean = PM25wMean,
            Year = 2011)
pm19NEavg <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  summarize(PM25Mean = mean(PM25_19, na.rm = TRUE),
            PM25wMean = weighted.mean(x = PM25_19,
                                      w = totalpopE, na.rm = TRUE)) %>% 
  transmute(STATE = "NEW ENGLAND", 
            PM25Mean = PM25Mean, 
            PM25wMean = PM25wMean,
            Year = 2016)
#rbind tables
PM11_16wSTAvg_actual <- bind_rows(PM15wSTAvgs_actual,
                                PM19wSTAvgs_actual,
                                pm15NEavg,pm19NEavg) %>% 
  mutate(Year = as.factor(Year),
         PM25Mean = round(PM25wMean,2),
         PM25wMean = round(PM25wMean,2))
# make slope graph
newggslopegraph(dataframe = PM11_16wSTAvg_actual, 
                Times = Year, 
                Measurement = PM25Mean, 
                Grouping = STATE,
                Title = expression(paste("Annual Average ", PM[2.5], " Concentrations")),
                SubTitle = expression(paste
                                      (PM[2.5]," (", mu, "g/", m^3, ")", sep = "")),
                Caption = NULL,
                LineColor = c("NEW ENGLAND" = "#000000",
                              "Connecticut" = "#E69F00",
                              "Massachusetts" = "#56B4E9",
                              "Rhode Island" = "#009E73",
                              "New Hampshire" = "#F0E442",
                              "Vermont" = "#0072B2",
                              "Maine" = "#D55E00"),
                LineThickness = 1,
                YTextSize = 3.5,
                DataTextSize = 3,
                WiderLabels = TRUE)
```


### PM~2.5~ in New England and Populations of Concern

In addition to variations in the general geography of PM~2.5~ concentrations, exposure to these pollutants also varies demographically. Figure \@ref(fig:barplotPM25PopAvgNE) below shows population-weighted exposures for populations of concern relative to average PM~2.5~ concentrations for the region. For example, minorities in New England are exposed to PM~2.5~ concentrations that are more than 5% above concentrations for the region as a whole. Similarly, linguistically isolated households are exposed to concentrations 4.9% above the regional average. By contrast, persons over age 64 are, on average, exposed to concentrations of PM~2.5~ at or below the regional average. 

```{r barplotPM25PopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Graph of population-weighted average exposures to PM2.5 for populations of concern in New England relative to the regional average."}
# Pop Weighted avg of PM2.5 for all Groups in New England relative to NE average
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(totalpopE,
                nhwhitepopE,
                minorityE,
                nhblackpopE,
                nhamerindpopE,
                nhasianpopE,
                nhnativhpopE,
                nhotherpopE,
                nh2morepopE,
                hisppopE,
                povknownE,
                num2povE, 
                eng_hhE,
                eng_limitE,
                age25upE,
                lthsE, 
                allAgesE, 
                under5E, 
                over64E, 
                PM25_19) %>% 
  gather(key = Group, value = Pop, totalpopE:over64E) %>% 
  group_by(Group) %>% 
  summarize(PM25wMean = weighted.mean(x = PM25_19, w = Pop, na.rm = TRUE),
            PM25Mean = mean(PM25_19, na.rm = TRUE)) %>% 
  spread(key = Group, value = PM25wMean) %>% 
  transmute(Minority = (minorityE/PM25Mean - 1)*100,
         #Minority_NHW = (minorityE/nhwhitepopE - 1)*100,
         `Lang Isol` = (eng_limitE/PM25Mean - 1)*100,
         `Low Income` = (num2povE/PM25Mean - 1)*100,
         `No HS` = (lthsE/PM25Mean - 1)*100,
         `Under 5` = (under5E/PM25Mean - 1)*100,
         `Over 64` = (over64E/PM25Mean - 1)*100) %>%
  gather(key = Group, value = Pct) %>% 
  ggplot(aes(x = reorder(Group, -Pct), y = Pct, fill = Group)) + 
  geom_bar(stat = "identity", position = "identity") +
  theme_minimal() +
  labs(x = "", y = "", title = expression(atop(paste("Population-Weighted ", PM[2.5], " Exposure"), "(relative to New England average)"))) + 
  theme(legend.position = 'none') +
  geom_text(aes(x = Group, y = Pct + 0.2 * sign(Pct), 
                label = paste0(round(Pct,2),"%")), 
            hjust = 0.5, size = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = 0)
```

Like the region as a whole, these populations have also experienced a decline in exposure since 2011. The comparison between exposure for these groups since 2011 is displayed below in Figure \@ref(fig:slopegraphPM25PopAvgNE). Note however that except for persons over age 64 and low income persons, all other populations of concern continue to experience exposures greater than the regional average, with minorities leading on this measure. 

```{r slopegraphPM25PopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Slope graph of annual PM2.5 concentrations by population of concern for 2011 and 2016."}
# Pop Weighted avg of PM2.5 EXPOSURE CHANGE 2011 - 2016 for all Groups in New England relative to NE average. 
# Create df of 2019 actual values
PM19wAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(minorityE,
                num2povE,
                eng_limitE,
                lthsE,
                under5E,
                over64E,
                PM25_19) %>% 
  gather(key = Group, value = Pop, minorityE:over64E) %>% 
  group_by(Group) %>% 
  summarize(PM25wMean = weighted.mean(x = PM25_19, 
                                      w = Pop, na.rm = TRUE)) %>% 
  mutate(Group = case_when(
    Group == "minorityE" ~ "Minority",
    Group == "eng_limitE" ~ "Lang Isol",
    Group == "num2povE" ~ "Low Income",
    Group == "lthsE" ~ "No HS Dip",
    Group == "under5E" ~ "Under 5",
    Group == "over64E" ~ "Over 64")) %>% 
  mutate(Year = 2016)
# Create df of 2015 values
PM15wAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(mins_15,
                lowinc_15,
                lingiso_15,
                lths_15,
                under5_15,
                over64_15,
                pm_15) %>% 
  gather(key = Group, value = Pop, mins_15:over64_15) %>% 
  group_by(Group) %>% 
  summarize(PM25wMean = weighted.mean(x = pm_15, 
                                      w = Pop, na.rm = TRUE)) %>%  
  mutate(Group = case_when(
    Group == "mins_15" ~ "Minority",
    Group == "lingiso_15" ~ "Lang Isol",
    Group == "lowinc_15" ~ "Low Income",
    Group == "lths_15" ~ "No HS Dip",
    Group == "under5_15" ~ "Under 5",
    Group == "over64_15" ~ "Over 64")) %>% 
  mutate(Year = 2011)
# Create regional average benchmark
pm15NEavg <- mean(ne_blkgrp_sf$pm_15, na.rm = TRUE)
pm19NEavg <- mean(ne_blkgrp_sf$PM25_19, na.rm = TRUE)
pmRegionalAvg_actual <- data.frame(
  Group = c("REGIONAL AVG", "REGIONAL AVG"),
  PM25wMean = c(pm15NEavg,pm19NEavg),
  Year = c(2011,2016)
)
#rbind tables
PM11_16wAvg_actual <- bind_rows(PM15wAvgs_actual,
                                PM19wAvgs_actual,
                                pmRegionalAvg_actual) %>% 
  mutate(Year = as.factor(Year),
         PM25wMean = round(PM25wMean,1))
# make slope graph
newggslopegraph(dataframe = PM11_16wAvg_actual, 
                Times = Year, 
                Measurement = PM25wMean, 
                Grouping = Group,
                Title = "Population-Weighted PM2.5 Exposure",
                SubTitle = expression(paste
                                      (PM[2.5]," (", mu, "g/", m^3, ")", sep = "")),
                Caption = NULL,
                LineColor = c("REGIONAL AVG" = "#000000",
                              "Lang Isol" = "#E69F00",
                              "Minority" = "#56B4E9",
                              "No HS" = "#009E73",
                              "Poverty" = "#F0E442",
                              "Under 5" = "#0072B2",
                              "Over 64" = "#D55E00"),
                YTextSize = 3.5,
                DataTextSize = 3,
                WiderLabels = TRUE)
```

Indeed, there is a moderate but statistically significant positive relationship between the proportion of minority residents or language-isolated households and the concentration of PM~2.5~ (see Figure \@ref(fig:cormatrixPM25PopAvg) below.).

```{r cormatrixPM25PopAvg, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Spearman's correlation matrix of annual PM2.5 concentrations and the proportions of populations of concern by Census Block Group."}
# Create a correlation matrix of PM25 and populations of concern
corrPM25 <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  dplyr::transmute(PM25 = PM25_19, 
                Minority = minority_pctE, 
                `Low Income` = pct2povE, 
                `Lang Isol` = eng_limit_pctE, 
                `No HS Dip` = pct_lthsE, 
                `Under 5` = pct_under5E, 
                `Over 64` = pct_over64E) %>% 
  drop_na() %>% 
  cor(method = "spearman")
# corrPM25
ggcorrplot(corrPM25, hc.order = TRUE, type = "lower", lab = TRUE, 
           title = "PM2.5 Correlation Matrix for New England", 
           legend.title = "Spearman's \nCorrelation\nCoefficient")
```


## Analysis of Ozone (O~3~) in New England

Ozone (O~3~) levels vary significantly across New England, with highest concentrations in southwest Connecticut, and declining toward the north and east (see Figure \@ref(fig:mapO3NE) below). 

```{r mapO3NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of 2016 Ozone summer seasonal average of daily maximum 8-hour concentration in air in parts per billion across New England at Census Block Group level."}

# Map of O3 across New England
tm_shape(ne_blkgrp_sf, unit = "mi") + 
  tm_fill("OZONE_19", style = "quantile", 
          title = expression(paste
                             (O[3]," (ppb)", sep = "")),
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=1)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Summer Ozone \nConcentrations\n2016", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

Ozone (O~3~) concentrations exhibit spatial clustering of both hot spots (i.e. geographic clusters of high values) and cold spots (i.e. geographic clusters of lower values). The map below (Figure \@ref(fig:hotspotO3NE))shows statistically significant Ozone (O~3~) hot spots.

```{r hotspotO3NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Hot spot map of 2016 Ozone concentrations at Census Block Group level."}
# Get rid of empty geometries and NAs, and convert to spdf
empty_geo <- st_is_empty(ne_blkgrp_sf)
ne_blkgrp_sp_DemoEJ <- ne_blkgrp_sf[!empty_geo,] %>% 
  dplyr::select(GEOID,OZONE_19) %>% 
  # st_transform(., crs = 2163) %>% # convert to US National Atlas Equal Area
  na.omit() %>% 
  as_Spatial()
# Calculate Queen's case neighbors
neighborsQC <- poly2nb(ne_blkgrp_sp_DemoEJ, queen = TRUE)
# Compute neighbor weights
# spdep::set.ZeroPolicyOption(TRUE)
listw <- nb2listw(neighborsQC, style = "W", zero.policy = TRUE)
# compute Getis-Ord Gi statistic
local_g <- localG(ne_blkgrp_sp_DemoEJ$OZONE_19, listw)
local_g <- cbind(ne_blkgrp_sp_DemoEJ, as.matrix(local_g))
names(local_g)[3] <- "gstat"
# map the results
tm_shape(local_g, unit = "mi",) + 
  tm_fill("gstat",
          palette = "-RdBu", 
          style = "pretty",
          title = expression(paste("Getis-Ord ", G[i]^"*")),
          showNA = FALSE,
          midpoint = NA,
          labels = c("Significant Clusters","of Low Values", "No Significant Clustering", "Significant Clusters","of High Values")) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Hot Spot Map of \nOzone for\nNew England", frame = FALSE, main.title.size = 0.6,
            legend.position = c(.8,.2),
            legend.title.size = 0.7)
```

There are statistically significant hot spots of Ozone (O~3~) primarily around Stamford, Connecticut. A warm cluster encompasses of all Connecticut, Rhode Island, and the southwestern half of Massachusetts. 

These Ozone (O~3~) levels vary significantly by state (see Figure \@ref(fig:boxplotO3NE) below). Connecticut stands out with the highest median summer seasonal average Ozone (O~3~) concentrations among the New England states, while Maine has the lowest overall concentrations. The differences in median concentrations between the states are all statistically significant.  

```{r boxplotO3NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Boxplot of 2016 Ozone summer seasonal average of daily maximum 8-hour concentrations in air in parts per billion by state at Census Block Group level."}
# boxplot of PM2.5 by state for 2016
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  ggplot(aes(x = STATE, y = OZONE_19, fill = STATE)) + 
  geom_boxplot(notch = TRUE) + 
  ggtitle(expression(atop(paste("Ozone (", O[3], ") Summer Concentration"), "by New England state, 2016"))) +
  theme_minimal() +
  theme(legend.position = "none", axis.text=element_text(size=8)) + 
  xlab(NULL) + 
  ylab(expression(paste(O[3]," (ppb)", sep = ""))) +
  coord_flip()
```

Since 2011, Ozone (O~3~) levels have increased slightly across the region, on average by `r {round(mean(ne_blkgrp_sf$OZONE_pctChange, na.rm = TRUE),1)}`%. These have changes not been uniform (see Figure \@ref(fig:mapO3NEchange) below). The greatest declines, up to 10.5%, have been in the northern half of Maine and Vermont, northeastern Vermont, southwestern Massachusetts, and southern Rhode Island. By contrast, increases in summer ozone concentrations of up to 8.1% appear in southwestern Connecticut, from Stamford to Hartford and north into western Massachusetts around Springfield. Significant increases also occur along the western border of Vermont, especially in the Champlain Valley. Minor increaes appear also in the greater Boston metro region, and southern and central Maine west and north of Portland.

```{r mapO3NEchange, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of percent change in summer seasonal Ozone concentrations across New England between 2011 and 2016 at Census Block Group level."}
# Map of O3 change across New England
tm_shape(ne_blkgrp_sf, unit = "mi") + 
  tm_fill("OZONE_pctChange", palette = "-RdYlGn",
          style = "quantile", 
          title = "% Change",
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=1)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Change in summer Ozone \nConcentrations \n2011-2016", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

Averaged across the states, the differences in these changes are also apparent. Figure \@ref(fig:slopegraphO3NE) below compares the average summer Ozone (O~3~) concentrations by state and for the region between 2011 and 2016. New England as a whole showed a slightly less than a 1% increase in summer average Ozone concentrations, while Connecticut showed an increase of nearly 5%. Massachusetts increased as well, but by only 0.5%. All other states showed an average decline in Ozone, with Rhode Island leading (-2.6%). Although note that Rhode Island's concentrations still remain significantly above the regional average. 

```{r slopegraphO3NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Slope graph of summer average Ozone concentrations for 2011 and 2016 for New England and states."}
# Slope graph of O3 by state and region between 2011 and 2016
# Create df of 2019 actual values
O319wSTAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(O3Mean = mean(OZONE_19, na.rm = TRUE),
            O3wMean = weighted.mean(x = OZONE_19,
                                      w = totalpopE, na.rm = TRUE)) %>% 
  mutate(Year = 2016)
# Create df of 2015 values
O315wSTAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(O3Mean = mean(o3_15, na.rm = TRUE),
            O3wMean = weighted.mean(x = o3_15,
                                      w = pop_15, na.rm = TRUE)) %>% 
  mutate(Year = 2011)
# Create regional average benchmark
O315NEavg <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  summarize(O3Mean = mean(o3_15, na.rm = TRUE),
            O3wMean = weighted.mean(x = o3_15,
                                      w = pop_15, na.rm = TRUE)) %>% 
  transmute(STATE = "NEW ENGLAND", 
            O3Mean = O3Mean, 
            O3wMean = O3wMean,
            Year = 2011)
O319NEavg <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  summarize(O3Mean = mean(OZONE_19, na.rm = TRUE),
            O3wMean = weighted.mean(x = OZONE_19,
                                      w = totalpopE, na.rm = TRUE)) %>% 
  transmute(STATE = "NEW ENGLAND", 
            O3Mean = O3Mean, 
            O3wMean = O3wMean,
            Year = 2016)
#rbind tables
O311_16wSTAvg_actual <- bind_rows(O315wSTAvgs_actual,
                                O319wSTAvgs_actual,
                                O315NEavg,
                                O319NEavg) %>% 
  mutate(Year = as.factor(Year),
         O3Mean = round(O3wMean,2),
         O3wMean = round(O3wMean,2))
# make slope graph
newggslopegraph(dataframe = O311_16wSTAvg_actual, 
                Times = Year, 
                Measurement = O3Mean, 
                Grouping = STATE,
                Title = expression(paste("Summer Average Ozone (", O[3], ") Concentrations")),
                SubTitle = expression(paste
                                      (O[3]," (ppb)", sep = "")),
                Caption = NULL,
                LineColor = c("NEW ENGLAND" = "#000000",
                              "Connecticut" = "#E69F00",
                              "Massachusetts" = "#56B4E9",
                              "Rhode Island" = "#009E73",
                              "New Hampshire" = "#F0E442",
                              "Vermont" = "#0072B2",
                              "Maine" = "#D55E00"),
                LineThickness = 1,
                YTextSize = 3.5,
                DataTextSize = 3,
                WiderLabels = TRUE)
```


### Ozone (O~3~) in New England and Populations of Concern

In addition to variations in the general geography of Ozone (O~3~) concentrations, exposure to this pollutant also varies demographically. Figure \@ref(fig:barplotO3PopAvgNE) below shows population-weighted exposures for populations of concern relative to average Ozone concentrations for the region. For example, minorities in New England are exposed to summer Ozone concentrations that are more than 3% above concentrations for the region as a whole. Similarly, linguistically isolated households are exposed to concentrations 2.6% above the regional average. By contrast, low income persons are, on average, exposed to concentrations of Ozone at or below the regional average. 

```{r barplotO3PopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Graph of population-weighted average exposures to Ozone for populations of concern in New England relative to the regional average."}
# Pop Weighted avg of O3 for all Groups in New England relative to NE average
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(totalpopE,
                nhwhitepopE,
                minorityE,
                nhblackpopE,
                nhamerindpopE,
                nhasianpopE,
                nhnativhpopE,
                nhotherpopE,
                nh2morepopE,
                hisppopE,
                povknownE,
                num2povE, 
                eng_hhE,
                eng_limitE,
                age25upE,
                lthsE, 
                allAgesE, 
                under5E, 
                over64E, 
                OZONE_19) %>% 
  gather(key = Group, value = Pop, totalpopE:over64E) %>% 
  group_by(Group) %>% 
  summarize(O3wMean = weighted.mean(x = OZONE_19, w = Pop, na.rm = TRUE),
            O3Mean = mean(OZONE_19, na.rm = TRUE)) %>% 
  spread(key = Group, value = O3wMean) %>% 
  transmute(Minority = (minorityE/O3Mean - 1)*100,
         #Minority_NHW = (minorityE/nhwhitepopE - 1)*100,
         `Lang Isol` = (eng_limitE/O3Mean - 1)*100,
         `Low Income` = (num2povE/O3Mean - 1)*100,
         `No HS` = (lthsE/O3Mean - 1)*100,
         `Under 5` = (under5E/O3Mean - 1)*100,
         `Over 64` = (over64E/O3Mean - 1)*100) %>%
  gather(key = Group, value = Pct) %>% 
  ggplot(aes(x = reorder(Group, -Pct), y = Pct, fill = Group)) + 
  geom_bar(stat = "identity", position = "identity") +
  theme_minimal() +
  labs(x = "", y = "", title = expression(atop(paste("Population-Weighted Ozone (", O[3], ") Exposure"), "(relative to New England average)"))) + 
  theme(legend.position = 'none') +
  geom_text(aes(x = Group, y = Pct + 0.2 * sign(Pct), 
                label = paste0(round(Pct,2),"%")), 
            hjust = 0.5, size = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = 0)
```

Like the region as a whole, these populations have also experienced changes in exposure since 2011. The comparison between exposure for these groups since 2011 is displayed below in Figure \@ref(fig:slopegraphO3PopAvgNE). All populations of concern have experienced an increase in population-weighted exposure to Ozone, with minorities and persons in language-isolated households experiencing the greatest increase in exposure. 

```{r slopegraphO3PopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Slope graph of summer Ozone concentrations by population of concern for 2011 and 2016."}
# Pop Weighted avg of O3 EXPOSURE CHANGE 2011 - 2016 for all Groups in New England relative to NE average. 
# Create df of 2019 actual values
O319wAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(minorityE,
                num2povE,
                eng_limitE,
                lthsE,
                under5E,
                over64E,
                OZONE_19) %>% 
  gather(key = Group, value = Pop, minorityE:over64E) %>% 
  group_by(Group) %>% 
  summarize(O3wMean = weighted.mean(x = OZONE_19, 
                                      w = Pop, na.rm = TRUE)) %>% 
  mutate(Group = case_when(
    Group == "minorityE" ~ "Minority",
    Group == "eng_limitE" ~ "Lang Isol",
    Group == "num2povE" ~ "Low Income",
    Group == "lthsE" ~ "No HS Dip",
    Group == "under5E" ~ "Under 5",
    Group == "over64E" ~ "Over 64")) %>% 
  mutate(Year = 2016)
# Create df of 2015 values
O315wAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(mins_15,
                lowinc_15,
                lingiso_15,
                lths_15,
                under5_15,
                over64_15,
                o3_15) %>% 
  gather(key = Group, value = Pop, mins_15:over64_15) %>% 
  group_by(Group) %>% 
  summarize(O3wMean = weighted.mean(x = o3_15, 
                                      w = Pop, na.rm = TRUE)) %>%  
  mutate(Group = case_when(
    Group == "mins_15" ~ "Minority",
    Group == "lingiso_15" ~ "Lang Isol",
    Group == "lowinc_15" ~ "Low Income",
    Group == "lths_15" ~ "No HS Dip",
    Group == "under5_15" ~ "Under 5",
    Group == "over64_15" ~ "Over 64")) %>% 
  mutate(Year = 2011)
# Create regional average benchmark
O315NEavg <- mean(ne_blkgrp_sf$o3_15, na.rm = TRUE)
O319NEavg <- mean(ne_blkgrp_sf$OZONE_19, na.rm = TRUE)
O3RegionalAvg_actual <- data.frame(
  Group = c("REGIONAL AVG", "REGIONAL AVG"),
  O3wMean = c(O315NEavg,O319NEavg),
  Year = c(2011,2016)
)
#rbind tables
O311_16wAvg_actual <- bind_rows(O315wAvgs_actual,
                                O319wAvgs_actual,
                                O3RegionalAvg_actual) %>% 
  mutate(Year = as.factor(Year),
         O3wMean = round(O3wMean,1))
# make slope graph
newggslopegraph(dataframe = O311_16wAvg_actual, 
                Times = Year, 
                Measurement = O3wMean, 
                Grouping = Group,
                Title = "Population-Weighted Ozone Exposure",
                SubTitle = expression(paste
                                      (O[3]," (ppb)", sep = "")),
                Caption = NULL,
                LineColor = c("REGIONAL AVG" = "#000000",
                              "Lang Isol" = "#E69F00",
                              "Minority" = "#56B4E9",
                              "No HS Dip" = "#009E73",
                              "Low Income" = "#F0E442",
                              "Under 5" = "#0072B2",
                              "Over 64" = "#D55E00"),
                YTextSize = 3.5,
                DataTextSize = 3,
                WiderLabels = TRUE)
```

Indeed, there is a moderate but statistically significant positive relationship between the proportion of minority residents and the concentration of Ozone (see Figure \@ref(fig:cormatrixO3PopAvg) below.).

```{r cormatrixO3PopAvg, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Spearman's correlation matrix of summer Ozone concentrations and the proportions of populations of concern by Census Block Group."}
# Create a correlation matrix of PM25 and populations of concern
corrO3 <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  dplyr::transmute(Ozone = OZONE_19, 
                Minority = minority_pctE, 
                `Low Income` = pct2povE, 
                `Lang Isol` = eng_limit_pctE, 
                `No HS Dip` = pct_lthsE, 
                `Under 5` = pct_under5E, 
                `Over 64` = pct_over64E) %>% 
  drop_na() %>% 
  cor(method = "spearman")
# corrPM25
ggcorrplot(corrO3, hc.order = TRUE, type = "lower", lab = TRUE, 
           title = "Ozone Correlation Matrix for New England", 
           legend.title = "Spearman's \nCorrelation\nCoefficient")
```



## Analysis of Carbon Dioxide (CO~2~) in New England

CO~2~ emissions closely follow major roadways arcoss the region (see Figure \@ref(fig:mapCO2NE) below). 

```{r mapCO2NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of 2017 Carbon Dioxide (CO2) annual on-road emissions in metric tons per square kilometer across New England at Census Block Group level."}

# Map of CO2 across New England
ne_blkgrp_sf %>% 
  mutate(onroad_2017 = kgco2_2017/1000/(bg_area_m2/1000000)) %>% 
  tm_shape(., unit = "mi") + 
  tm_fill("onroad_2017", style = "quantile", 
          title = expression(paste
                             (CO[2]," (mtons/", km^2,")", sep = "")),
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=0)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Annual On-road \nCarbon Dioxide\nEmissions, 2017", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

Significant hot spots, or clusters of high CO~2~ emissions, appear around major New England cities, including Hartford, Providence, Portsmouth, Portland, and along the I-495 belt around Boston (see Figure \@ref(fig:hotspotCO2NE) below).

```{r hotspotCO2NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Hot spot map of 2017 CO2 emissions at Census Block Group level."}
# Get rid of empty geometries and NAs, and convert to spdf
empty_geo <- st_is_empty(ne_blkgrp_sf)
ne_blkgrp_sp_DemoEJ <- ne_blkgrp_sf[!empty_geo,] %>% 
  mutate(onroad_2017 = kgco2_2017/1000) %>%
  dplyr::select(GEOID,onroad_2017) %>% 
  # st_transform(., crs = 2163) %>% # convert to US National Atlas Equal Area
  na.omit() %>% 
  as_Spatial()
# Calculate Queen's case neighbors
neighborsQC <- poly2nb(ne_blkgrp_sp_DemoEJ, queen = TRUE)
# Compute neighbor weights
# spdep::set.ZeroPolicyOption(TRUE)
listw <- nb2listw(neighborsQC, style = "W", zero.policy = TRUE)
# compute Getis-Ord Gi statistic
local_g <- localG(ne_blkgrp_sp_DemoEJ$onroad_2017, listw)
local_g <- cbind(ne_blkgrp_sp_DemoEJ, as.matrix(local_g))
names(local_g)[3] <- "gstat"
# map the results
tm_shape(local_g, unit = "mi",) + 
  tm_fill("gstat",
          palette = "-RdBu", 
          style = "pretty",
          title = expression(paste("Getis-Ord ", G[i]^"*")),
          showNA = FALSE,
          midpoint = NA,
          labels = c("Significant Clusters","of Low Values", "Significant Clusters","of High Values")) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Hot Spot Map of \nCO2 for\nNew England", frame = FALSE, main.title.size = 0.6,
            legend.position = c(.8,.2),
            legend.title.size = 0.7)
```

CO~2~ emissions vary by state (see Table \@ref(tab:statsCO2NE) and Figure \@ref(fig:boxplotCO2NE) below). Massachusetts shows highest absolute emissions.  

```{r statsCO2NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Annual 2017 CO2 emissions (mtons) at Census Block Group level and by state."}
ne_blkgrp_sf %>% 
  mutate(onroad_2017 = kgco2_2017/1000,
         onroad_perkm_2017 = kgco2_2017/1000/(bg_area_m2/1000000)) %>%
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(Mean = round(mean(onroad_perkm_2017, na.rm = TRUE),0),
            Median = round(median(onroad_perkm_2017, na.rm = TRUE),0),
            Min = round(min(onroad_perkm_2017, na.rm = TRUE),0),
            Max = round(max(onroad_perkm_2017, na.rm = TRUE),0),
            `State Total` = sum(onroad_2017, na.rm = TRUE)) %>% 
  kableExtra::kable(format.args = list(big.mark = ','), 
                    caption = "Annual 2017 CO~2~ emissions (mtons/km^2^) by Census block group and total CO~2~ emissions (mtons) by state")
```


```{r boxplotCO2NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Boxplot of annual on-road Carbon Dioxoide emissions in metric tons per square kilometer by state at Census Block Group level."}
# boxplot of CO2 by state for 2017
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  mutate(onroad_2017 = kgco2_2017/1000/(bg_area_m2/1000000)) %>%
  ggplot(aes(x = STATE, y = onroad_2017, fill = STATE)) + 
  geom_boxplot(notch = TRUE) + 
  ggtitle(expression(atop(paste("Carbon Dioxide (", CO[2], ") Emissions"), "by Census block group by New England state, 2017"))) +
  theme_minimal() +
  theme(legend.position = "none", axis.text=element_text(size=8)) + 
  xlab(NULL) + 
  ylab(expression(paste(CO[2]," (mtons/", km^2, ")", sep = ""))) +
  coord_flip()
```

Since 1990, CO~2~ emissions increased significantly across the region, on average by `r {round((sum(ne_blkgrp_sf$kgco2_2017, na.rm = TRUE)-sum(ne_blkgrp_sf$kgco2_1990, na.rm = TRUE))/sum(ne_blkgrp_sf$kgco2_1990, na.rm = TRUE)*100,1)}`%. These changes have not been uniform (see Figure \@ref(fig:mapCO2NEchange) below). Except for Maine, the greatest declines, up to 97%, have been in outer suburban and rural areas. By contrast, increases in CO~2~ emissions of over 600% appear across most urban cores. Boston is the notable exception where increases have been relatively modest. Dramtic increases characterize areas around Worcester, Massachusetts and the South Shore of Massachusetts and the Cape, as well as around Portland, Maine, and from Stamford to Hartford, Connecticut. Significant increases also occur west and north of Montpelier, Vermont.

```{r mapCO2NEchange, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of percent change in annual CO2 emissions across New England between 1990 and 2017 at Census Block Group level."}
# Map of CO2 change across New England
ne_blkgrp_sf %>% 
  mutate(CO2_pctChange90_17 = (kgco2_2017 - kgco2_1990)/kgco2_1990*100) %>%
  tm_shape(., unit = "mi") + 
  tm_fill("CO2_pctChange90_17", palette = "-RdYlGn",
          style = "quantile", 
          title = "% Change",
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=1)),
          legend.is.portrait = TRUE,
          midpoint = 0) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Change in annual CO2 \nEmissions \n1990-2017", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

Averaged across the states, the differences in these changes are also apparent, although less extreme than at the Census block group level. Table \@ref(tab:changeCO2NE) and Figure \@ref(fig:slopegraphCO2NE) below compares the  annual CO~2~ emissions by state and for the region between 1990 and 2017. New England as a whole showed an approximately 20% increase in annual CO~2~ emissions. All states showed total increases, with New Hampshire leading with a 34% increase, followed by Maine with a 26% increase. Rhode Island showed the smallest increase since 1990 (2%), although note that Rhode Island's average annual emissions still remain significantly above the regional average. 

```{r changeCO2NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Annual 2017 CO2 emissions (mtons) and Percent Change by state at Census Block Group level."}
# acquire table of 1990 state pop values
library(tidycensus)
census_api_key("f2776fbc29cf847505de9308a82c8d65290d16b3")
ne_states_df90 <- map_df(ne_states, function(x) {
    get_decennial(geography = "state", 
            variables = c(pop90 = "P0010001"),
            state = x, output = "wide", year = 1990)
  }) %>% 
  select(-GEOID)

ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  mutate(onroad_2017 = kgco2_2017/1000,
         onroad_1990 = kgco2_1990/1000) %>%
  group_by(STATE) %>% 
  summarize(`1990 CO2 (mtons)` = sum(onroad_1990, na.rm = TRUE),
            `2017 CO2 (mtons)` = sum(onroad_2017, na.rm = TRUE),
            `Pct Change` = 
              paste0(
                round(
                  (`2017 CO2 (mtons)` - `1990 CO2 (mtons)`)/`1990 CO2 (mtons)`*100,0),"%")) %>% 
  left_join(., ne_states_df, by = c("STATE" = "NAME")) %>% 
  mutate(`2017 Per Capita (mtons/person)` = 
              `2017 CO2 (mtons)`/totalpopE) %>% 
  left_join(., ne_states_df90, by = c("STATE" = "NAME")) %>% 
  mutate(`1990 Per Capita (mtons/person)` = `1990 CO2 (mtons)`/pop90,
         `Per Capita Pct Change` = 
           paste0(
           round(
             (`2017 Per Capita (mtons/person)` - `1990 Per Capita (mtons/person)`)/`1990 Per Capita (mtons/person)`*100,0),"%")) %>% 
  transmute(STATE = STATE, 
            `1990 CO2 (mtons)` = `1990 CO2 (mtons)`, 
            `2017 CO2 (mtons)` = `2017 CO2 (mtons)`,
            `Change` = `Pct Change`,
            `1990 Per Capita` = 
              as.character(round(`1990 Per Capita (mtons/person)`,2)),
            `2017 Per Capita` = 
              as.character(round(`2017 Per Capita (mtons/person)`,2)),
            `Per Capita Change` = `Per Capita Pct Change`) %>% 
  kableExtra::kable(longtable = T, booktabs = T,
                    format.args = list(big.mark = ','),
                    caption = "Annual On-road Carbon Dioxide Emissions",
                    digits = 0, align = "r") %>% 
  kableExtra::kable_styling(latex_options = c("repeat_header"))
```


```{r slopegraphCO2NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Slope graph of total CO2 emissions for 1990 and 2017 for New England and states."}
# Slope graph of CO2 by state and region between 1990 and 2017
# Create df of 2017 actual values
CO217ST_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(CO2total = sum(kgco2_2017, na.rm = TRUE)/1000) %>% 
  mutate(Year = 2017)
# Create df of 1990 values
CO290ST_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(CO2total = sum(kgco2_1990, na.rm = TRUE)/1000) %>% 
  mutate(Year = 1990)
# Create regional average benchmark
CO290NE <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  summarize(CO2total = sum(kgco2_1990, na.rm = TRUE)/1000) %>% 
  transmute(STATE = "NEW ENGLAND", 
            CO2total = CO2total,
            Year = 1990)
CO217NE <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  summarize(CO2total = sum(kgco2_2017, na.rm = TRUE)/1000) %>% 
  transmute(STATE = "NEW ENGLAND", 
            CO2total = CO2total,
            Year = 2017)
#rbind tables
CO290_17ST_actual <- bind_rows(CO290ST_actual,
                                CO217ST_actual,
                                CO290NE,
                                CO217NE) %>% 
  mutate(Year = as.factor(Year),
         CO2total = round(CO2total,0))
# make slope graph
newggslopegraph(dataframe = CO290_17ST_actual, 
                Times = Year, 
                Measurement = CO2total, 
                Grouping = STATE,
                Title = expression(paste("Annual On-road ", 
                                         CO[2], " Emissions")),
                SubTitle = expression(paste
                                      (CO[2]," (mtons)",  sep = "")),
                Caption = NULL,
                LineColor = c("NEW ENGLAND" = "#000000",
                              "Connecticut" = "#E69F00",
                              "Massachusetts" = "#56B4E9",
                              "Rhode Island" = "#009E73",
                              "New Hampshire" = "#F0E442",
                              "Vermont" = "#0072B2",
                              "Maine" = "#D55E00"),
                LineThickness = 1,
                YTextSize = 3.5,
                DataTextSize = 3,
                WiderLabels = TRUE)
```

The growth in CO~2~ emissions exceeds population growth across the region and in most states as is evident in the per capita emissions (see last three columns in Table \@ref(tab:changeCO2NE)). While New Hampshire leads in percentage growth of total emissions for the region, Maine leads in per capita growth. Only Vermont and Rhode Island showed declines in per capita emissions. 

### Carbon Dioxide (CO~2~) in New England and Populations of Concern

In addition to variations in the general geography of CO~2~ emissions, exposure to these emissions also varies demographically. Figure \@ref(fig:barplotO3PopAvgNE) below shows population-weighted exposures for populations of concern relative to average CO~2~ emissions per square kilometer for the region. For example, language-isolated households in New England are exposed to  CO~2~ emissions per square kilometer that are more than 69% above emissions for the region as a whole. Similarly, minorities are exposed to emissions over 49% above the regional average. By contrast, persons in over age 64 are, on average, exposed to emissions of CO~2~ over 14% below the regional average. 

```{r barplotCO2PopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Graph of population-weighted average exposures to Carbon Dioxide emissions per square kilometer for populations of concern in New England relative to the regional average."}
# Pop Weighted avg of CO2 for all Groups in New England relative to NE average
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  mutate(co2_2017 = kgco2_2017/1000/(bg_area_m2/1000000)) %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(totalpopE,
                nhwhitepopE,
                minorityE,
                nhblackpopE,
                nhamerindpopE,
                nhasianpopE,
                nhnativhpopE,
                nhotherpopE,
                nh2morepopE,
                hisppopE,
                povknownE,
                num2povE, 
                eng_hhE,
                eng_limitE,
                age25upE,
                lthsE, 
                allAgesE, 
                under5E, 
                over64E, 
                co2_2017) %>% 
  gather(key = Group, value = Pop, totalpopE:over64E) %>% 
  group_by(Group) %>% 
  summarize(CO2wMean = weighted.mean(x = co2_2017, w = Pop, na.rm = TRUE),
            CO2Mean = mean(co2_2017, na.rm = TRUE)) %>% 
  spread(key = Group, value = CO2wMean) %>% 
  transmute(Minority = (minorityE/CO2Mean - 1)*100,
         #Minority_NHW = (minorityE/nhwhitepopE - 1)*100,
         `Lang Isol` = (eng_limitE/CO2Mean - 1)*100,
         `Low Income` = (num2povE/CO2Mean - 1)*100,
         `No HS` = (lthsE/CO2Mean - 1)*100,
         `Under 5` = (under5E/CO2Mean - 1)*100,
         `Over 64` = (over64E/CO2Mean - 1)*100) %>%
  gather(key = Group, value = Pct) %>% 
  ggplot(aes(x = reorder(Group, -Pct), y = Pct, fill = Group)) + 
  geom_bar(stat = "identity", position = "identity") +
  theme_minimal() +
  labs(x = "", y = "", title = expression(atop(paste("Population-Weighted ", CO[2], " Exposure"), "(relative to New England average)"))) + 
  theme(legend.position = 'none') +
  geom_text(aes(x = Group, y = Pct + 0.2 * sign(Pct), 
                label = paste0(round(Pct,1),"%")), 
            hjust = 0.5, size = 3, nudge_y = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = 0)
```

Like the region as a whole, these populations have also experienced changes in exposure since 1990. The comparison between exposure for these groups since 1990 is displayed below in Figure \@ref(fig:slopegraphCO2PopAvgNE). All populations of concern have experienced an increase in population-weighted exposure to CO~2~, with persons in over age 64 and minorities  experiencing the greatest increase in exposure. 

```{r slopegraphCO2PopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Slope graph of carbon dioxide emissions by population of concern for 1990 and 2017."}
# Pop Weighted avg of CO2 EXPOSURE CHANGE 1990 - 2017 for all Groups in New England relative to NE average. 
# Create df of 2017 actual values
CO217wAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  mutate(mtco2_2017 = kgco2_2017/1000/(bg_area_m2/1000000)) %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(minorityE,
                num2povE,
                eng_limitE,
                lthsE,
                under5E,
                over64E,
                mtco2_2017) %>% 
  gather(key = Group, value = Pop, minorityE:over64E) %>% 
  group_by(Group) %>% 
  summarize(CO2wMean = weighted.mean(x = mtco2_2017, 
                                      w = Pop, na.rm = TRUE)) %>% 
  mutate(Group = case_when(
    Group == "minorityE" ~ "Minority",
    Group == "eng_limitE" ~ "Lang Isol",
    Group == "num2povE" ~ "Low Income",
    Group == "lthsE" ~ "No HS Dip",
    Group == "under5E" ~ "Under 5",
    Group == "over64E" ~ "Over 64")) %>% 
  mutate(Year = 2017)
# Create df of 1990 values
CO290wAvgs_actual <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  select(GEOID,bg_area_m2) %>% 
  left_join(ne_blkgrp_sf90,., by = c("STFID" = "GEOID")) %>% 
  as.data.frame() %>% 
  mutate(mtco2_1990 = kgco2_1990/1000/(bg_area_m2/1000000)) %>%
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(minority,
                lowInc,
                langIsolHH,
                noHSdip,
                under5,
                over64,
                mtco2_1990) %>% 
  drop_na() %>% 
  gather(key = Group, value = Pop, minority:over64) %>% 
  group_by(Group) %>% 
  summarize(CO2wMean = weighted.mean(x = mtco2_1990, 
                                      w = Pop, na.rm = TRUE)) %>%  
  mutate(Group = case_when(
    Group == "minority" ~ "Minority",
    Group == "langIsolHH" ~ "Lang Isol",
    Group == "lowInc" ~ "Low Income",
    Group == "noHSdip" ~ "No HS Dip",
    Group == "under5" ~ "Under 5",
    Group == "over64" ~ "Over 64")) %>% 
  mutate(Year = 1990)
# Create regional average benchmark
CO290NEavg <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  mutate(co2_1990 = kgco2_1990/1000/(bg_area_m2/1000000)) %>% 
  summarize(mean(co2_1990, na.rm = TRUE)) %>% 
  pull()
CO217NEavg <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  mutate(co2_2017 = kgco2_2017/1000/(bg_area_m2/1000000)) %>% 
  summarize(mean(co2_2017, na.rm = TRUE)) %>% 
  pull()
CO2RegionalAvg_actual <- data.frame(
  Group = c("REGIONAL AVG", "REGIONAL AVG"),
  CO2wMean = c(CO290NEavg,CO217NEavg),
  Year = c(1990,2017)
)
#rbind tables
CO290_17wAvg_actual <- bind_rows(CO290wAvgs_actual,
                                CO217wAvgs_actual,
                                CO2RegionalAvg_actual) %>% 
  mutate(Year = as.factor(Year),
         CO2wMean = round(CO2wMean,1))
# make slope graph
newggslopegraph(dataframe = CO290_17wAvg_actual, 
                Times = Year, 
                Measurement = CO2wMean, 
                Grouping = Group,
                Title = "Annual Population-Weighted CO2 Exposure",
                SubTitle = expression(paste
                                      (CO[2]," (mtons/", km^2,")", sep = "")),
                Caption = NULL,
                LineColor = c("REGIONAL AVG" = "#000000",
                              "Lang Isol" = "#E69F00",
                              "Minority" = "#56B4E9",
                              "No HS Dip" = "#009E73",
                              "Low Income" = "#F0E442",
                              "Under 5" = "#0072B2",
                              "Over 64" = "#D55E00"),
                YTextSize = 3.5,
                DataTextSize = 3,
                WiderLabels = TRUE)
```

All populations of concern, except for persons over age 64, are more exposed to CO~2~ emissions. There is a moderately positive correlation between the proportions of minorities and language isolated households and emissions of CO~2~ per square kilometer (see Figure \@ref(fig:cormatrixCO2PopAvg) below.).

```{r cormatrixCO2PopAvg, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Spearman's correlation matrix of annual average carbon dioxide emissions and the proportions of populations of concern by Census Block Group."}
# Create a correlation matrix of PM25 and populations of concern
corrCO2 <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  dplyr::transmute(CO2 = kgco2_2017/1000/(bg_area_m2/1000000), 
                Minority = minority_pctE, 
                `Low Income` = pct2povE, 
                `Lang Isol` = eng_limit_pctE, 
                `No HS Dip` = pct_lthsE, 
                `Under 5` = pct_under5E, 
                `Over 64` = pct_over64E) %>% 
  drop_na() %>% 
  cor(method = "spearman")
# corrPM25
ggcorrplot(corrCO2, hc.order = TRUE, type = "lower", lab = TRUE, 
           title = "CO2 Correlation Matrix for New England", 
           legend.title = "Spearman's \nCorrelation\nCoefficient",
           ggtheme = ggplot2::theme_minimal())
```



## Analysis of Diesel Particulate Matter in New England

Diesel Particulate Matter (PM) emissions are concentrated around major cities in New England and radiate outward along major traffic routes (see Figure \@ref(fig:mapDPMNE) below). 

```{r mapDPMNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of 2014 annual average ambient concentrations of Diesel Particulate Matter in micrograms per cubic meter across New England at Census Block Group level."}

# Map of DPM across New England
ne_blkgrp_sf %>% 
  tm_shape(., unit = "mi") + 
  tm_fill("DSLPM_19", style = "quantile", 
          title = expression(paste
                             ("DPM"," (", mu, "g/", m^3, ")", sep = "")),
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=3)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Annual Average \nDiesel PM Emissions\n2014", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

There appear to be no significant spatial clusters of high Diesel Paticulate Matter concentrations (see Figure \@ref(fig:hotspotDPMNE) below).

```{r hotspotDPMNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Hot spot map of 2014 Diesel Particulate Matter emissions at Census Block Group level."}
# Get rid of empty geometries and NAs, and convert to spdf
empty_geo <- st_is_empty(ne_blkgrp_sf)
ne_blkgrp_sp_DemoEJ <- ne_blkgrp_sf[!empty_geo,] %>%
  dplyr::select(GEOID,DSLPM_19) %>% 
  # st_transform(., crs = 2163) %>% # convert to US National Atlas Equal Area
  na.omit() %>% 
  as_Spatial()
# Calculate Queen's case neighbors
neighborsQC <- poly2nb(ne_blkgrp_sp_DemoEJ, queen = TRUE)
# Compute neighbor weights
# spdep::set.ZeroPolicyOption(TRUE)
listw <- nb2listw(neighborsQC, style = "W", zero.policy = TRUE)
# compute Getis-Ord Gi statistic
local_g <- localG(ne_blkgrp_sp_DemoEJ$DSLPM_19, listw)
local_g <- cbind(ne_blkgrp_sp_DemoEJ, as.matrix(local_g))
names(local_g)[3] <- "gstat"
# map the results
tm_shape(local_g, unit = "mi",) + 
  tm_fill("gstat",
          palette = "-RdBu", 
          style = "pretty",
          title = expression(paste("Getis-Ord ", G[i]^"*")),
          showNA = FALSE,
          midpoint = NA,
          labels = c("Significant Clusters","of Low Values", "No Significant Clusters", "Significant Clusters","of High Values")) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Hot Spot Map of \nDPM for\nNew England", frame = FALSE, main.title.size = 0.6,
            legend.position = c(.8,.2),
            legend.title.size = 0.7)
```

Diesel Particulate Matter concentrations vary by state (see Table \@ref(tab:statsDPMNE) and Figure \@ref(fig:boxplotDPMNE) below). Rhode Island shows the highest average ambient concentrations, although Massachusetts shows more extreme values.  

```{r statsDPMNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Annual 2014 average ambient concentrations of Diesel Particulate Matter in micrograms per cubic meter across New England at Census Block Group level."}
ne_blkgrp_sf %>%
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(Mean = round(mean(DSLPM_19, na.rm = TRUE),3),
            Median = round(median(DSLPM_19, na.rm = TRUE),3),
            Min = round(min(DSLPM_19, na.rm = TRUE),3),
            Max = round(max(DSLPM_19, na.rm = TRUE),3)) %>% 
  kableExtra::kable(format.args = list(big.mark = ','), 
                    caption = "Annual 2014 DPM concentrations by Census block group")
```


```{r boxplotDPMNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Boxplot of 7 Carbon Dioxoide on-road emissions in metric tons per square kilometer by state at Census Block Group level."}
# boxplot of CO2 by state for 2017
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  ggplot(aes(x = STATE, y = DSLPM_19, fill = STATE)) + 
  geom_boxplot(notch = TRUE) + 
  ggtitle("Diesel Particulate Matter (DPM) concentrations \nby Census block group by New England state, 2017") +
  theme_minimal() +
  theme(legend.position = "none", axis.text=element_text(size=8)) + 
  xlab(NULL) + 
  ylab(expression(paste("DPM"," (", mu, "g/", m^3, ")", sep = ""))) +
  coord_flip()
```


### Diesel Particulate Matter in New England and Populations of Concern

In addition to variations in the general geography of Diesel Particulate Matter concentrations, exposure to these concentrations also varies demographically. Figure \@ref(fig:barplotDPMPopAvgNE) below shows population-weighted exposures for populations of concern relative to average Diesel Particulate Matter concentraions for the region. For example, linguistically isolated households in New England are exposed to  Diesel Particulate Matter concentrations that are more than 45% above concentrations for the region as a whole. Similarly, minorities are exposed to concentrations more than 30% above the regional average. By contrast, persons over age 64 are, on average, exposed to concentrations of Diesel Particulate Matter 9% below the regional average. 

```{r barplotDPMPopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Graph of population-weighted average exposures to annual average ambient concentrations of Diesel Particulate Matter across New England relative to the regional average."}
# Pop Weighted avg of DPM for all Groups in New England relative to NE average
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(totalpopE,
                nhwhitepopE,
                minorityE,
                nhblackpopE,
                nhamerindpopE,
                nhasianpopE,
                nhnativhpopE,
                nhotherpopE,
                nh2morepopE,
                hisppopE,
                povknownE,
                num2povE, 
                eng_hhE,
                eng_limitE,
                age25upE,
                lthsE, 
                allAgesE, 
                under5E, 
                over64E, 
                DSLPM_19) %>% 
  gather(key = Group, value = Pop, totalpopE:over64E) %>% 
  group_by(Group) %>% 
  summarize(DPMwMean = weighted.mean(x = DSLPM_19, w = Pop, na.rm = TRUE),
            DPMMean = mean(DSLPM_19, na.rm = TRUE)) %>% 
  spread(key = Group, value = DPMwMean) %>% 
  transmute(Minority = (minorityE/DPMMean - 1)*100,
         #Minority_NHW = (minorityE/nhwhitepopE - 1)*100,
         `Lang Isol` = (eng_limitE/DPMMean - 1)*100,
         `Low Income` = (num2povE/DPMMean - 1)*100,
         `No HS` = (lthsE/DPMMean - 1)*100,
         `Under 5` = (under5E/DPMMean - 1)*100,
         `Over 64` = (over64E/DPMMean - 1)*100) %>%
  gather(key = Group, value = Pct) %>% 
  ggplot(aes(x = reorder(Group, -Pct), y = Pct, fill = Group)) + 
  geom_bar(stat = "identity", position = "identity") +
  theme_minimal() +
  labs(x = "", y = "", title = "Population-Weighted Exposure to DPM \n(relative to New England average)") + 
  theme(legend.position = 'none') +
  geom_text(aes(x = Group, y = Pct + 0.2 * sign(Pct), 
                label = paste0(round(Pct,1),"%")), 
            hjust = 0.5, size = 3, nudge_y = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = 0)
```

Language isolated households and minorities are much more exposed to Diesel Particulate Matter concentrations than the general population. There is a strong positive correlation between the proportions of minorities and ambient concentrations of Diesel Particulate Matter, and a moderately strong positive correlation with language isolated households (see Figure \@ref(fig:cormatrixDPMPopAvg) below.).

```{r cormatrixDPMPopAvg, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Spearman's correlation matrix of annual average Diesel Particulate Matter concentrations and the proportions of populations of concern by Census Block Group."}
# Create a correlation matrix of PM25 and populations of concern
corrDPM <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  dplyr::transmute(DPM = DSLPM_19, 
                Minority = minority_pctE, 
                `Low Income` = pct2povE, 
                `Lang Isol` = eng_limit_pctE, 
                `No HS Dip` = pct_lthsE, 
                `Under 5` = pct_under5E, 
                `Over 64` = pct_over64E) %>% 
  drop_na() %>% 
  cor(method = "spearman")
# corrPM25
ggcorrplot(corrDPM, hc.order = TRUE, type = "lower", lab = TRUE, 
           title = "DPM Correlation Matrix for New England", 
           legend.title = "Spearman's \nCorrelation\nCoefficient",
           ggtheme = ggplot2::theme_minimal())
```



## Analysis of Air Toxics Cancer Risk in New England

Air Toxics Cancer Risk are concentrated around major cities in New England and radiate outward along major traffic routes (see Figure \@ref(fig:mapATCNE) below). 

```{r mapATCNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of 2014 lifetime cancer risk from inhalation of air toxics (expressed as risk in-1 million) across New England at Census Block Group level."}

# Map of DPM across New England
ne_blkgrp_sf %>% 
  tm_shape(., unit = "mi") + 
  tm_fill("CANCER_19", style = "quantile", 
          title = "Risk-in-1 million",
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=1)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Annual Lifetime Cancer Risk \nfrom Inhalation of Air Toxics\n2014", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

There appear to be significant spatial clusters of high lifetime cancer risk from inhalation of air toxics in Boston and small areas in southwestern Connecticut (see Figure \@ref(fig:hotspotATCNE) below).

```{r hotspotATCNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Hot spot map of 2014 lifetime cancer risk from inhalation of air toxics at Census Block Group level."}
# Get rid of empty geometries and NAs, and convert to spdf
empty_geo <- st_is_empty(ne_blkgrp_sf)
ne_blkgrp_sp_DemoEJ <- ne_blkgrp_sf[!empty_geo,] %>%
  dplyr::select(GEOID,CANCER_19) %>% 
  # st_transform(., crs = 2163) %>% # convert to US National Atlas Equal Area
  na.omit() %>% 
  as_Spatial()
# Calculate Queen's case neighbors
neighborsQC <- poly2nb(ne_blkgrp_sp_DemoEJ, queen = TRUE)
# Compute neighbor weights
# spdep::set.ZeroPolicyOption(TRUE)
listw <- nb2listw(neighborsQC, style = "W", zero.policy = TRUE)
# compute Getis-Ord Gi statistic
local_g <- localG(ne_blkgrp_sp_DemoEJ$CANCER_19, listw)
local_g <- cbind(ne_blkgrp_sp_DemoEJ, as.matrix(local_g))
names(local_g)[3] <- "gstat"
# map the results
tm_shape(local_g, unit = "mi",) + 
  tm_fill("gstat",
          palette = "-RdBu", 
          style = "pretty",
          title = expression(paste("Getis-Ord ", G[i]^"*")),
          showNA = FALSE,
          midpoint = NA,
          labels = c("Significant Clusters","of Low Values", "No Significant Clusters", "Significant Clusters","of High Values")) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Hot Spot Map of \nAir Toxics Cancer Risk for\nNew England", frame = FALSE, main.title.size = 0.6,
            legend.position = c(.8,.2),
            legend.title.size = 0.7)
```

Lifetime cancer risk from inhalation of air toxics vary by state (see Table \@ref(tab:statsATCNE) and Figure \@ref(fig:boxplotATCNE) below). Massachusetts, Connecticut, and Rhode Island show the highest average lifetime cancer risks.  

```{r statsATCNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Annual 2014 lifetime cancer risk from inhalation of air toxics (expressed as risk in-1 million) across New England at Census Block Group level."}
ne_blkgrp_sf %>%
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(Mean = round(mean(CANCER_19, na.rm = TRUE),1),
            Median = round(median(CANCER_19, na.rm = TRUE),1),
            Min = round(min(CANCER_19, na.rm = TRUE),1),
            Max = round(max(CANCER_19, na.rm = TRUE),1)) %>% 
  kableExtra::kable(format.args = list(big.mark = ','), 
                    caption = "Annual 2014 Cancer Risk from Inhalation of Air Toxics by Census block group (risk in-1 million)")
```


```{r boxplotATCNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Boxplot of lifetime cancer risk from inhalation of air toxics (expressed as risk in-1 million) across New England by state at Census Block Group level."}
# boxplot of lifetime cancer risk by state for 2014
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  ggplot(aes(x = STATE, y = CANCER_19, fill = STATE)) + 
  geom_boxplot(notch = TRUE) + 
  ggtitle("Lifetime Cancer Risk from \nInhalation of Air Toxics \nby Census block group by New England state, 2017") +
  theme_minimal() +
  theme(legend.position = "none", axis.text=element_text(size=8)) + 
  xlab(NULL) + 
  ylab("risk in-1 million") +
  coord_flip()
```


### Lifetime Cancer Risk from Inhalation of Air Toxics in New England and Populations of Concern

In addition to variations in the general geography of Diesel Particulate Matter concentrations, exposure to these concentrations also varies demographically. Figure \@ref(fig:barplotATCPopAvgNE) below shows population-weighted exposures for populations of concern relative to average Lifetime Cancer Risk from Inhalation of Air Toxics for the region. For example, linguistically isolated households in New England experience lifetime cancer risks from inhalation of air toxics that are 11% above the region as a whole. Similarly, minorities are experience lifetime cancer risks of 9% above the regional average. By contrast, persons over age 64 are, on average, exposed to concentrations of Diesel Particulate Matter 2% below the regional average. 

```{r barplotATCPopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Graph of population-weighted average lifetime cancer risk from inhalation of air toxics (expressed as risk in-1 million) across New England at Census Block Group level relative to the regional average."}
# Pop Weighted avg of DPM for all Groups in New England relative to NE average
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(totalpopE,
                nhwhitepopE,
                minorityE,
                nhblackpopE,
                nhamerindpopE,
                nhasianpopE,
                nhnativhpopE,
                nhotherpopE,
                nh2morepopE,
                hisppopE,
                povknownE,
                num2povE, 
                eng_hhE,
                eng_limitE,
                age25upE,
                lthsE, 
                allAgesE, 
                under5E, 
                over64E, 
                CANCER_19) %>% 
  gather(key = Group, value = Pop, totalpopE:over64E) %>% 
  group_by(Group) %>% 
  summarize(DPMwMean = weighted.mean(x = CANCER_19, w = Pop, na.rm = TRUE),
            DPMMean = mean(CANCER_19, na.rm = TRUE)) %>% 
  spread(key = Group, value = DPMwMean) %>% 
  transmute(Minority = (minorityE/DPMMean - 1)*100,
         #Minority_NHW = (minorityE/nhwhitepopE - 1)*100,
         `Lang Isol` = (eng_limitE/DPMMean - 1)*100,
         `Low Income` = (num2povE/DPMMean - 1)*100,
         `No HS` = (lthsE/DPMMean - 1)*100,
         `Under 5` = (under5E/DPMMean - 1)*100,
         `Over 64` = (over64E/DPMMean - 1)*100) %>%
  gather(key = Group, value = Pct) %>% 
  ggplot(aes(x = reorder(Group, -Pct), y = Pct, fill = Group)) + 
  geom_bar(stat = "identity", position = "identity") +
  theme_minimal() +
  labs(x = "", y = "", title = "Population-Weighted Average Lifetime Cancer Risk \n(relative to New England average)") + 
  theme(legend.position = 'none') +
  geom_text(aes(x = Group, y = Pct + 0.2 * sign(Pct), 
                label = paste0(round(Pct,1),"%")), 
            hjust = 0.5, size = 3, nudge_y = 0.5,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = 0)
```

Language isolated households and minorities experience higher lifetime risk of cancer from inhalation of air toxics than the general population. There is a moderate to strong positive correlation between the proportions of minorities and cancer risk, and a moderately strong positive correlation with language isolated households (see Figure \@ref(fig:cormatrixATCPopAvg) below.).

```{r cormatrixATCPopAvg, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Spearman's correlation matrix of lifetime cancer risk and the proportions of populations of concern by Census Block Group."}
# Create a correlation matrix of cancer risk and populations of concern
corrACT <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  dplyr::transmute(`Cancer Risk` = CANCER_19, 
                Minority = minority_pctE, 
                `Low Income` = pct2povE, 
                `Lang Isol` = eng_limit_pctE, 
                `No HS Dip` = pct_lthsE, 
                `Under 5` = pct_under5E, 
                `Over 64` = pct_over64E) %>% 
  drop_na() %>% 
  cor(method = "spearman")

ggcorrplot(corrACT, hc.order = TRUE, type = "lower", lab = TRUE, 
           title = "Cancer Risk Correlation Matrix for New England", 
           legend.title = "Spearman's \nCorrelation\nCoefficient",
           ggtheme = ggplot2::theme_minimal())
```



## Analysis of Respiratory Hazard Index in New England

Higher Respiratory Hazard Indices are concentrated around major cities in New England and radiate outward along major traffic routes (see Figure \@ref(fig:mapRHINE) below). 

```{r mapRHINE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of 2014 respiratory hazard index from inhalation of air toxics (expressed as ratio of exposure concentration to health-based reference concentration) across New England at Census Block Group level."}

# Map of RHI across New England
ne_blkgrp_sf %>% 
  tm_shape(., unit = "mi") + 
  tm_fill("RESP_19", style = "quantile", 
          title = "Ratio of exposure to \nhealth-based reference \nconcentration",
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=1)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Annual Respiratory Hazard \nIndex from Inhalation of \nAir Toxics\n2014", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

There appear to be significant spatial clusters of high lifetime cancer risk from inhalation of air toxics in Boston (see Figure \@ref(fig:hotspotRHINE) below).

```{r hotspotRHINE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Hot spot map of 2014 respiratory hazard index from inhalation of air toxics at Census Block Group level."}
# Get rid of empty geometries and NAs, and convert to spdf
empty_geo <- st_is_empty(ne_blkgrp_sf)
ne_blkgrp_sp_DemoEJ <- ne_blkgrp_sf[!empty_geo,] %>%
  dplyr::select(GEOID,RESP_19) %>% 
  # st_transform(., crs = 2163) %>% # convert to US National Atlas Equal Area
  na.omit() %>% 
  as_Spatial()
# Calculate Queen's case neighbors
neighborsQC <- poly2nb(ne_blkgrp_sp_DemoEJ, queen = TRUE)
# Compute neighbor weights
# spdep::set.ZeroPolicyOption(TRUE)
listw <- nb2listw(neighborsQC, style = "W", zero.policy = TRUE)
# compute Getis-Ord Gi statistic
local_g <- localG(ne_blkgrp_sp_DemoEJ$RESP_19, listw)
local_g <- cbind(ne_blkgrp_sp_DemoEJ, as.matrix(local_g))
names(local_g)[3] <- "gstat"
# map the results
tm_shape(local_g, unit = "mi",) + 
  tm_fill("gstat",
          palette = "-RdBu", 
          style = "pretty",
          title = expression(paste("Getis-Ord ", G[i]^"*")),
          showNA = FALSE,
          midpoint = NA,
          labels = c("Significant Clusters","of Low Values", "", "","Significant Clusters","of High Values")) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Hot Spot Map of \nRespiratory \nHazard Index \nfor New England", frame = FALSE, main.title.size = 0.6,
            legend.position = c(.8,.2),
            legend.title.size = 0.7)
```

Lifetime cancer risk from inhalation of air toxics vary by state (see Table \@ref(tab:statsRHINE) and Figure \@ref(fig:boxplotRHINE) below). Massachusetts, Connecticut, and Rhode Island show the highest average lifetime respiratory hazard indices.  

```{r statsRHINE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Annual 2014 respiratory hazard index from inhalation of air toxics (expressed as ratio of exposure concentration to health-based reference concentration) across New England at Census Block Group level."}
ne_blkgrp_sf %>%
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(Mean = round(mean(RESP_19, na.rm = TRUE),2),
            Median = round(median(RESP_19, na.rm = TRUE),2),
            Min = round(min(RESP_19, na.rm = TRUE),2),
            Max = round(max(RESP_19, na.rm = TRUE),2)) %>% 
  kableExtra::kable(format.args = list(big.mark = ','), 
                    caption = "Annual 2014 Respiratory Hazard Index from Inhalation of Air Toxics by Census block group (ratio of exposure concentration to health-based reference concentration)")
```


```{r boxplotRHINE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Boxplot of respiratory hazard index from inhalation of air toxics (expressed as ratio of exposure concentration to health-based reference concentration) across New England by state at Census Block Group level."}
# boxplot of lifetime cancer risk by state for 2014
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  ggplot(aes(x = STATE, y = RESP_19, fill = STATE)) + 
  geom_boxplot(notch = TRUE) + 
  ggtitle("Respiratory Hazard Index from Inhalation of \nAir Toxics by Census block group \nby New England state, 2017") +
  theme_minimal() +
  theme(legend.position = "none", axis.text=element_text(size=8)) + 
  xlab(NULL) + 
  ylab("ratio of exposure concentration to health-based reference concentration") +
  coord_flip()
```


### Respiratory Hazard Index from Inhalation of Air Toxics in New England and Populations of Concern

In addition to variations in the general geography of the Respitory Hazard Index, exposure to these risks also varies demographically. Figure \@ref(fig:barplotRHIPopAvgNE) below shows population-weighted exposures for populations of concern relative to average Respiratory Hazard Index from Inhalation of Air Toxics for the region. For example, linguistically isolated households in New England experience lifetime cancer risks from inhalation of air toxics that are 15% above the region as a whole. Similarly, minorities are experience lifetime hazard index values more than 11% above the regional average. By contrast, persons over age 64 are, on average, exposed to respiratory hazard indices over 2% below the regional average. 

```{r barplotRHIPopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Graph of population-weighted average respiratory hazard index from inhalation of air toxics (expressed as ratio of exposure concentration to health-based reference concentration) across New England at Census Block Group level relative to the regional average."}
# Pop Weighted avg of RHI for all Groups in New England relative to NE average
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(totalpopE,
                nhwhitepopE,
                minorityE,
                nhblackpopE,
                nhamerindpopE,
                nhasianpopE,
                nhnativhpopE,
                nhotherpopE,
                nh2morepopE,
                hisppopE,
                povknownE,
                num2povE, 
                eng_hhE,
                eng_limitE,
                age25upE,
                lthsE, 
                allAgesE, 
                under5E, 
                over64E, 
                RESP_19) %>% 
  gather(key = Group, value = Pop, totalpopE:over64E) %>% 
  group_by(Group) %>% 
  summarize(DPMwMean = weighted.mean(x = RESP_19, w = Pop, na.rm = TRUE),
            DPMMean = mean(RESP_19, na.rm = TRUE)) %>% 
  spread(key = Group, value = DPMwMean) %>% 
  transmute(Minority = (minorityE/DPMMean - 1)*100,
         #Minority_NHW = (minorityE/nhwhitepopE - 1)*100,
         `Lang Isol` = (eng_limitE/DPMMean - 1)*100,
         `Low Income` = (num2povE/DPMMean - 1)*100,
         `No HS` = (lthsE/DPMMean - 1)*100,
         `Under 5` = (under5E/DPMMean - 1)*100,
         `Over 64` = (over64E/DPMMean - 1)*100) %>%
  gather(key = Group, value = Pct) %>% 
  ggplot(aes(x = reorder(Group, -Pct), y = Pct, fill = Group)) + 
  geom_bar(stat = "identity", position = "identity") +
  theme_minimal() +
  labs(x = "", y = "", title = "Population-Weighted Average Respiratory Hazard Index \n(relative to New England average)") + 
  theme(legend.position = 'none') +
  geom_text(aes(x = Group, y = Pct + 0.2 * sign(Pct), 
                label = paste0(round(Pct,1),"%")), 
            hjust = 0.5, size = 3, nudge_y = 0.5,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = 0)
```

Language isolated households and minorities experience higher respiratory hazard indices from inhalation of air toxics compared to the general population. There is a moderate to strong positive correlation between the proportions of minorities and respiratory hazard index, and a moderately strong positive correlation with language isolated households (see Figure \@ref(fig:cormatrixRHIPopAvg) below.).

```{r cormatrixRHIPopAvg, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Spearman's correlation matrix of lifetime cancer risk and the proportions of populations of concern by Census Block Group."}
# Create a correlation matrix of cancer risk and populations of concern
corrACT <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  dplyr::transmute(`Resp Hazard Index` = RESP_19, 
                Minority = minority_pctE, 
                `Low Income` = pct2povE, 
                `Lang Isol` = eng_limit_pctE, 
                `No HS Dip` = pct_lthsE, 
                `Under 5` = pct_under5E, 
                `Over 64` = pct_over64E) %>% 
  drop_na() %>% 
  cor(method = "spearman")
# corrPM25
ggcorrplot(corrACT, hc.order = TRUE, type = "lower", lab = TRUE, 
           title = "Respiratory Hazard Index \nCorrelation Matrix for New England", 
           legend.title = "Spearman's \nCorrelation\nCoefficient",
           ggtheme = ggplot2::theme_minimal())
```



## Analysis of Traffic Proximity and Volume in New England

Traffic Proximity and Volume is a count of vehicles (average annual daily traffic) at major roads within 500 meters, divided by distance in kilometers (km). It was calculated from 2017 U.S. DOT traffic data and the data for this analysis was sourced from EPAs EJSCREEN database. Exposure to high annual daily traffic volume is concentrated around major cities in New England and radiates outward along major traffic routes (see Figure \@ref(fig:mapTPVNE) below). 

```{r mapTPVNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of 2017 traffic proximity and volume (calculated as a count of vehicles (average annual daily traffic) at major roads within 500 meters, divided by distance in kilometers (km)) across New England at Census Block Group level."}

# Map of TPV across New England
ne_blkgrp_sf %>% 
  tm_shape(., unit = "mi") + 
  tm_fill("PTRAF_19", style = "quantile", 
          title = "AADT/Distance(km)",
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=1)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Annual Traffic \nProximity and Volume \n2017", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

There appear to be significant spatial clusters of traffic proximity and volume in Boston (see Figure \@ref(fig:hotspotTPVNE) below).

```{r hotspotTPVNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Hot spot map of 2017 traffic proximity and volume at Census Block Group level."}
# Get rid of empty geometries and NAs, and convert to spdf
empty_geo <- st_is_empty(ne_blkgrp_sf)
ne_blkgrp_sp_DemoEJ <- ne_blkgrp_sf[!empty_geo,] %>%
  dplyr::select(GEOID,PTRAF_19) %>% 
  # st_transform(., crs = 2163) %>% # convert to US National Atlas Equal Area
  na.omit() %>% 
  as_Spatial()
# Calculate Queen's case neighbors
neighborsQC <- poly2nb(ne_blkgrp_sp_DemoEJ, queen = TRUE)
# Compute neighbor weights
# spdep::set.ZeroPolicyOption(TRUE)
listw <- nb2listw(neighborsQC, style = "W", zero.policy = TRUE)
# compute Getis-Ord Gi statistic
local_g <- localG(ne_blkgrp_sp_DemoEJ$PTRAF_19, listw)
local_g <- cbind(ne_blkgrp_sp_DemoEJ, as.matrix(local_g))
names(local_g)[3] <- "gstat"
# map the results
tm_shape(local_g, unit = "mi",) + 
  tm_fill("gstat",
          palette = "-RdBu", 
          style = "pretty",
          title = expression(paste("Getis-Ord ", G[i]^"*")),
          showNA = FALSE,
          midpoint = NA,
          labels = c("Significant Clusters","of Low Values", "","No Significant Clusters", "","Significant Clusters","of High Values")) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Hot Spot Map of \nTraffic Proximity \nand Volume \nfor New England", frame = FALSE, main.title.size = 0.6,
            legend.position = c(.8,.2),
            legend.title.size = 0.7)
```

Traffic Proximity and Volume exposure vary by state (see Table \@ref(tab:statsTPVNE) and Figure \@ref(fig:boxplotTPVNE) below). Massachusetts shows the highest Traffic Proximity and Volume exposure, followed distantly by Connecticut, and Rhode Island.  

```{r statsTPVNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Annual 2017 traffic proximity and volume (calculated as a count of vehicles (average annual daily traffic) at major roads within 500 meters, divided by distance in kilometers (km)) across New England at Census Block Group level."}
ne_blkgrp_sf %>%
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(Mean = round(mean(PTRAF_19, na.rm = TRUE),0),
            Median = round(median(PTRAF_19, na.rm = TRUE),0),
            Min = round(min(PTRAF_19, na.rm = TRUE),0),
            Max = round(max(PTRAF_19, na.rm = TRUE),0)) %>% 
  kableExtra::kable(format.args = list(big.mark = ','), 
                    caption = "Annual traffic proximity and volume (AADT/Distance(km))")
```


```{r boxplotTPVNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Boxplot of 2017 traffic proximity and volume (calculated as a count of vehicles (average annual daily traffic) at major roads within 500 meters, divided by distance in kilometers (km)) across New England at Census Block Group level."}
# boxplot of traffic proximity and volume by state for 2017
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  ggplot(aes(x = STATE, y = PTRAF_19, fill = STATE)) + 
  geom_boxplot(notch = TRUE) + 
  ggtitle("Traffic Proximity and Volume \nby Census block group \nby New England state, 2017") +
  theme_minimal() +
  theme(legend.position = "none", axis.text=element_text(size=8)) + 
  xlab(NULL) + 
  ylab("AADT/Distance(km)") +
  coord_flip()
```


### Traffic Proximity and Volume Exposure in New England and Populations of Concern

In addition to variations in the general geography of the Traffic Proximity and Volume exposure, this exposure also varies demographically. Figure \@ref(fig:barplotTPVPopAvgNE) below shows population-weighted exposures for populations of concern relative to average Traffic Proximity and Volume for the region. For example, linguistically isolated households in New England are exposed to Traffic Proximity and Volume at more than twice the level of the region as a whole (over 104%). Similarly, minorities are exposed to Traffic Proximity and Volume more than 57% above the regional average. By contrast, persons over age 64 are, on average, are exposed to Traffic Proximity and Volume of more than 22% below the regional average. 

```{r barplotTPVPopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Graph of population-weighted average Traffic Proximity and Volume (calculated as a count of vehicles (average annual daily traffic) at major roads within 500 meters, divided by distance in kilometers (km)) across New England at Census Block Group level relative to the regional average."}
# Pop Weighted avg of TPV for all Groups in New England relative to NE average
ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(totalpopE,
                nhwhitepopE,
                minorityE,
                nhblackpopE,
                nhamerindpopE,
                nhasianpopE,
                nhnativhpopE,
                nhotherpopE,
                nh2morepopE,
                hisppopE,
                povknownE,
                num2povE, 
                eng_hhE,
                eng_limitE,
                age25upE,
                lthsE, 
                allAgesE, 
                under5E, 
                over64E, 
                PTRAF_19) %>% 
  gather(key = Group, value = Pop, totalpopE:over64E) %>% 
  group_by(Group) %>% 
  summarize(DPMwMean = weighted.mean(x = PTRAF_19, w = Pop, na.rm = TRUE),
            DPMMean = mean(PTRAF_19, na.rm = TRUE)) %>% 
  spread(key = Group, value = DPMwMean) %>% 
  transmute(Minority = (minorityE/DPMMean - 1)*100,
         #Minority_NHW = (minorityE/nhwhitepopE - 1)*100,
         `Lang Isol` = (eng_limitE/DPMMean - 1)*100,
         `Low Income` = (num2povE/DPMMean - 1)*100,
         `No HS` = (lthsE/DPMMean - 1)*100,
         `Under 5` = (under5E/DPMMean - 1)*100,
         `Over 64` = (over64E/DPMMean - 1)*100) %>%
  gather(key = Group, value = Pct) %>% 
  ggplot(aes(x = reorder(Group, -Pct), y = Pct, fill = Group)) + 
  geom_bar(stat = "identity", position = "identity") +
  theme_minimal() +
  labs(x = "", y = "", title = "Population-Weighted Average Traffic Proximity and Volume \n(relative to New England average)") + 
  theme(legend.position = 'none') +
  geom_text(aes(x = Group, y = Pct + 0.2 * sign(Pct), 
                label = paste0(round(Pct,1),"%")), 
            hjust = 0.5, size = 3, nudge_y = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = 0)
```

Language isolated households and minorities experience higher exposure to traffic proximity and volume compared to the general population. There is a moderate positive correlation between the proportions of minorities and traffic proximity and volume, and a moderate positive correlation with language isolated households (see Figure \@ref(fig:cormatrixTPVPopAvg) below.).

```{r cormatrixTPVPopAvg, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Spearman's correlation matrix of lifetime cancer risk and the proportions of populations of concern by Census Block Group."}
# Create a correlation matrix of cancer risk and populations of concern
corrTPV <- ne_blkgrp_sf %>% 
  as.data.frame() %>% 
  dplyr::transmute(`Traffic Exposure` = PTRAF_19, 
                Minority = minority_pctE, 
                `Low Income` = pct2povE, 
                `Lang Isol` = eng_limit_pctE, 
                `No HS Dip` = pct_lthsE, 
                `Under 5` = pct_under5E, 
                `Over 64` = pct_over64E) %>% 
  drop_na() %>% 
  cor(method = "spearman")
# corrPM25
ggcorrplot(corrTPV, hc.order = TRUE, type = "lower", lab = TRUE, 
           title = "Traffic Proximity and Volume \nCorrelation Matrix for New England", 
           legend.title = "Spearman's \nCorrelation\nCoefficient",
           ggtheme = ggplot2::theme_minimal())
```
