---
title: "Evacuation Risks in RI"
author: "Marcos Luna and Neenah Estrella-Luna"
date: "3/14/2020"
output: 
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: yes        
    includes:  
      in_header: my_header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak
# Analysis of evacuation-related risks in Rhode Island

This is an analysis of flood or storm surge-related evacuation risks for populations of concern in Rhode Island. In the event of significant inland or coastal flooding, evacuation may be required. For individuals and households with limited mobility, either due to inadequate access to transportation options or because of physical limitations, evacuation presents heightened risk. Evacuation may also prove especially difficult for individuals and households due to limited economic resources, difficulty understanding or accessing information, or low trust in official sources of information. 

## Analysis of Flood Hazards in Rhode Island

```{r NFHZAdata, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Analysis of flooding and hurricane evacuation risks for populations of concern in Rhode Island
library(tidyverse)
library(sf)
library(tmap)
library(tmaptools)
library(lwgeom)
library(tigris)
options(tigris_use_cache = TRUE, tigris_class = "sf")

load("DATA/ne_layers.rds")

# Convert to projected local CRS EPSG:2840: NAD83(HARN) / Rhode Island
ri_blkgrp_2840 <- ne_blkgrp_sf %>% 
  filter(STATE == "Rhode Island") %>% 
  st_transform(., crs = 2840)

ri_tracts_2840 <- ne_tracts_sf %>% 
  filter(STATE == "Rhode Island") %>% 
  st_transform(., crs = 2840)

# Get rid of empty geometries
empty_geo <- st_is_empty(ri_blkgrp_2840)
ri_blkgrp_2840 <- ri_blkgrp_2840[!empty_geo,]
empty_geo <- st_is_empty(ri_tracts_2840)
ri_tracts_2840 <- ri_tracts_2840[!empty_geo,]

# Download Census TIGERLine hydrography for RI
## First, extract list of county names to use with tigris::water
ri_counties <- counties("RI") %>% 
  pull(NAME)
# Next, download water features for each county and rbind to one layer
ri_awater_sf <- rbind_tigris(
  lapply(
    ri_counties, function(x) area_water(state = "RI", county = x)
  )
) %>% 
  st_union() %>% 
  st_as_sf() %>% 
  st_transform(., crs = 2840)

# Read in NFHL for RI. Data comes from FEMA. 
# List available layers in geodatabase
# st_layers("DATA/FEMA/RI/NFHL_44_20181118.gdb")
# Read in flood hazard areas
ri_fhza_2840 <- st_read(dsn = "DATA/FEMA/RI/NFHL_44_20181118.gdb", 
                        layer = "S_Fld_Haz_Ar") %>% 
  filter(FLD_ZONE != "OPEN WATER" & 
           !ZONE_SUBTY %in% c("AREA OF MINIMAL FLOOD HAZARD",
                              "AREA WITH REDUCED FLOOD RISK DUE TO LEVEE")) %>%
  mutate(FLD_ZONE = as.character(FLD_ZONE)) %>% # omit unused factor levels
  st_transform(., crs = 2840) %>% 
  st_make_valid() %>% 
  group_by(FLD_ZONE) %>% # aggregate flood zone polygons
  summarize(count = n()) %>% 
  mutate(Area = st_area(.),
         Interval = case_when(
           FLD_ZONE  == "A" ~ "100-year",
           FLD_ZONE  == "AE" ~ "100-year",
           FLD_ZONE  == "AH" ~ "100-year",
           FLD_ZONE  == "AO" ~ "100-year",
           FLD_ZONE  == "VE" ~ "100-year",
           FLD_ZONE == "X" ~ "500-year"))

# get rid of empty geometries
empty_geo <- st_is_empty(ri_fhza_2840)
ri_fhza_2840 <- ri_fhza_2840[!empty_geo,]

# crop flood zones to land areas only
ri_state_sf <- ne_states_sf_cb %>% 
  filter(NAME == "Rhode Island")
ri_fhza_2840_land <- ri_fhza_2840 %>% 
  crop_shape(., ri_state_sf, polygon = TRUE) %>% 
  st_difference(., ri_awater_sf) %>% 
  mutate(Area = st_area(.))

# read in processed ri_blkgrps and ri_tracts
# st_layers(dsn = "DATA/FEMA/RI")
ri_blkgrps_nfhza <- st_read(dsn = "DATA/FEMA/RI", 
                            layer = "ri_blkgrps_nfhza") %>% 
  left_join(., as.data.frame(ri_blkgrp_2840), by = "GEOID") %>% 
  st_transform(., crs = 2840) %>% 
  mutate(NewArea = st_area(.)) %>%
  st_make_valid()

ri_tracts_nfhza <- st_read(dsn = "DATA/FEMA/RI", 
                           layer = "ri_tracts_nfhza") %>% 
  left_join(., as.data.frame(ri_tracts_2840), by = "GEOID") %>% 
  st_transform(., crs = 2840) %>% 
  mutate(NewArea = st_area(.)) %>%
  st_make_valid()

# Apportion populations based on geographic proportion of intersect
ri_blkgrps_nfhza <- ri_blkgrps_nfhza %>%
  mutate(RI_LOWINC = if_else(RI_INCOME == "I",totalpopE,0)) %>%
  mutate(RI_LOWINC = replace_na(RI_LOWINC,0)) %>%
  mutate(RI_MINORITIES = if_else(RI_MINORITY == "M", totalpopE,0)) %>%
  mutate(RI_MINORITIES = replace_na(RI_MINORITIES,0)) %>%
  mutate(Proportion = as.numeric(NewArea/OldArea),
         NewPop = totalpopE*Proportion,
         NewMinority = minorityE*Proportion,
         NewUnder5 = under5E*Proportion,
         NewOver64 = over64E*Proportion,
         NewUnder18 = under18E*Proportion,
         NewEng_limit = eng_limitE*Proportion,
         NewPov = num2povE*Proportion,
         NewLths = lthsE*Proportion,
         NewRI_LOWINC = RI_LOWINC*Proportion,
         NewRI_MINORITIES = RI_MINORITIES*Proportion)

ri_tracts_nfhza <- ri_tracts_nfhza %>%
  mutate(Proportion = as.numeric(NewArea/OldArea),
         NewDisabled = disabledOver18E*Proportion,
         NewNoCar = HHnoCarE*Proportion)

# Compute total block group populations within flood zones
ri_flood_blkgrp_df <- ri_blkgrps_nfhza %>%
  as.data.frame() %>%
  summarize(`Total Pop` = as.integer(sum(NewPop)),
            Minority = as.integer(sum(NewMinority)),
            `Under 5` = as.integer(sum(NewUnder5)),
            `Over 64` = as.integer(sum(NewOver64)),
            `Under 18` = as.integer(sum(NewUnder18)),
            `Limited English HH` = as.integer(sum(NewEng_limit)),
            `Low Income` = as.integer(sum(NewPov)),
            `No HS Dip` = as.integer(sum(NewLths)),
            `RI Low Income` = as.integer(sum(NewRI_LOWINC)),
            `RI Minority` = as.integer(sum(NewRI_MINORITIES))) %>%
  gather(key = Group, value = FloodPop)

ri_flood_tracts_df <- ri_tracts_nfhza %>%
  as.data.frame() %>%
  summarize(`Disabled` = as.integer(sum(NewDisabled)),
            `No Car HH` = as.integer(sum(NewNoCar))) %>%
  gather(key = Group, value = FloodPop)

# Compute total tract populations within the state for same groups
ri_tract_flood_pops_df <- ri_tracts_2840 %>%
  as.data.frame() %>%
  summarize(`Disabled` = sum(disabledOver18E),
            `No Car HH` = sum(HHnoCarE)) %>%
  gather(key = Group, value = RIPop) %>%
  left_join(.,ri_flood_tracts_df, by = "Group")

# Compute populations for state,and join with flood pops
ri_FloodPops_df <- ri_blkgrp_2840 %>%
  as.data.frame() %>%
  mutate(RI_LOWINC = if_else(RI_INCOME == "I",totalpopE,0)) %>%
  mutate(RI_LOWINC = replace_na(RI_LOWINC,0)) %>%
  mutate(RI_MINORITIES = if_else(RI_MINORITY == "M",totalpopE,0)) %>%
  mutate(RI_MINORITIES = replace_na(RI_MINORITIES,0)) %>%
  summarize(`Total Pop` = sum(totalpopE),
            Minority = sum(minorityE),
            `Under 5` = sum(under5E),
            `Over 64` = sum(over64E),
            `Under 18` = sum(under18E),
            `Limited English HH` = sum(eng_limitE),
            `Low Income` = sum(num2povE),
            `No HS Dip` = sum(lthsE),
            `RI Low Income` = sum(RI_LOWINC, na.rm = TRUE),
            `RI Minority` = sum(RI_MINORITIES, na.rm = TRUE)) %>%
  gather(key = Group, value = RIPop) %>%
  left_join(., ri_flood_blkgrp_df, by = "Group") %>%
  rbind(.,ri_tract_flood_pops_df) %>%
  mutate(PctFlood = FloodPop/RIPop*100)
```

As a relatively flat, coastal plain state, a significant portion of Rhode Island's land area and population are exposed to the risk of flooding from overbanking of inland water bodies (e.g., ponds and rivers) or from coastal storm surge and sea level rise. 

The analysis of flood exposure presented here is based on the Federal Emergency Management Agency's (FEMA) National Flood Hazard Layer (NFHL), a digital version of FEMA's most recent flood maps.  FEMA identifies areas subject to varying levels of flood risk through its Flood Insurance Rate Maps (FIRMs), which are produced for most parts of the country to identify areas subject to flooding. FEMA's FIRMs are a national standard used by all federal agencies for the purposes of requiring and rating the purchase of flood insurance and regulating new development.  

Most of Rhode Island's population lives near the coast or near inland water bodies. As a result a significant proportion of the population lives near or within known flood zones. Figure \@ref(fig:mapNFHZAPop) below shows the population distribution and areas subject to floods with an Annual Exceedance Probability (AEP) of 1% (also known as a '100-year' flood) and areas subject to 0.2% AEP (also known as a '500-year' flood). Areas within the 1% AEP flood zone are designated by FEMA as Special Flood Hazard Areas, and development within those zones must be covered by flood insurance. Areas within the 0.2% AEP are not currently regulated, but these areas are nevertheless subject to flood risk under more extreme, albeit less frequent, flooding circumstances. Coastal areas within the 0.2% AEP zone are a reasonable proxy measure for the risks of sea level rise. Simulations of SLR and coastal inundation by NOAA show that areas covered by the 0.2% AEP are also areas that would be subject to the 1% AEP with SLR up to 7 feet above current mean sea level. 

```{r mapNFHZAPop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="FEMA flood zones and population distribution across Rhode Island at Census Block Group level."}
# Create a dot density map of total populations and overlay on flood zones
# Create point layer of major cities for context
# Note cb=FALSE is necessary for extracting centroids from town polygons. Otherwise, if cb=TRUE, cannot extract centroids from multipolygon features.
ri_towns_sf_pts <- county_subdivisions(state = "RI", cb = TRUE) %>% 
  filter(NAME %in% c("Providence", 
                     "Woonsocket", 
                     "Pawtucket", 
                     "Warwick", 
                     "Bristol", 
                     "Portsmouth", 
                     "Newport", 
                     "Narragansett", 
                     "North Kingstown", 
                     "Charlestown", 
                     "Situate",
                     "Glocester")) %>% 
  st_transform(., crs = 2840) %>% 
  st_centroid(of_largest_polygon = TRUE)

# Create road layer for context
ri_highways <- tigris::primary_roads() %>% 
  filter(FULLNAME %in% c("I- 95","I- 195","I- 295", "US Hwy 6")) %>%
  tmaptools::crop_shape(., ne_states_sf_cb) %>% 
  st_transform(., crs = 2840)

# Extract highway segments for labeling
I95roadSegment <- ri_highways %>% 
  filter(LINEARID == "110468245978")

I95roadSegment2 <- ri_highways %>% 
  filter(LINEARID == "1107052605232")

I295roadSegment <- ri_highways %>% 
  filter(LINEARID == "1104755623349")

I195roadSegment <- ri_highways %>% 
  filter(LINEARID == "110448466166")

# Create custom icons of highway shields
I95 <- tmap_icons(file = "https://upload.wikimedia.org/wikipedia/commons/thumb/6/61/I-95.svg/200px-I-95.svg.png")
I195 <- tmap_icons("https://upload.wikimedia.org/wikipedia/commons/thumb/f/f7/I-195.svg/200px-I-195.svg.png")
I295 <- tmap_icons("https://upload.wikimedia.org/wikipedia/commons/thumb/4/41/I-295.svg/200px-I-295.svg.png")

# Create random points, with 1 point for every 100 people
ri_totalpop_pts <- ri_blkgrp_2840 %>% 
  select(totalpopE) %>% 
  filter(totalpopE >= 100) %>% 
  st_sample(., size = round(.$totalpopE/100)) %>% # create 1 random point for every 100 people
  st_sf(.) %>% 
  mutate(Group = "Total Pop")

# Map totalpop and flood zones
tm_layout(bg.color = "#e6f3f7") +
  tm_shape(ri_blkgrp_2840, unit = "mi") + tm_fill(col = "white") +
  tm_shape(ne_states_sf_cb) + tm_fill(col="white") +
  tm_shape(ri_awater_sf) + tm_fill(col = "#e6f3f7") + 
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ri_totalpop_pts) + tm_dots(col = "forestgreen",
                                      labels = "1 dot = 100 persons") +
  tm_shape(ri_fhza_2840_land) + 
  tm_fill(col = "Interval", 
          palette = c("gold", "goldenrod3"), 
          labels = c("1% AEP (100-year)", "0.2% AEP (500-year)"), 
          title = "FEMA Flood Zones",
          alpha = 0.6,
          border.alpha = 0) +
  tm_shape(ri_fhza_2840_land) +
  tm_borders(col = "goldenrod1",
             lwd = 0.5,
             alpha = 0.6) +
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ri_highways) + tm_lines(col = "seashell4", lwd = 2) +
  tm_shape(I95roadSegment) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I95roadSegment2) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I195roadSegment) +
  tm_symbols(shape = I195, border.lwd = NA, size = 0.25) +
  tm_shape(I295roadSegment) + 
  tm_symbols(shape = I295, border.lwd = NA, size = 0.25) +
  tm_shape(ri_towns_sf_pts) + tm_dots(size = 1.5) +
  tm_text("NAME", size = 0.5, col = "black", 
          xmod = 0.7, ymod = 0.2, shadow = TRUE) +
  tm_scale_bar(breaks = c(0, 5, 10), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_add_legend(type = "fill", col = "forestgreen",
                border.col = "white", border.alpha = 0,
                labels = "1 dot = 100 people",
                title = "Total Population") +
  tm_layout(title = "Population Distribution \nand Flood Zones", 
            frame = TRUE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"))
```

Approximately 11.6% of Rhode Island's land area falls within FEMA flood zones. The breakdown by type of flood zone is presented in Table \@ref(tab:tabNFHZAarea) below. 

```{r tabNFHZAarea, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Rhode Island Land Area within Flood Zones."}
# Percentage of RI land within flood zones
ri_area <- as.numeric(ri_state_sf$ALAND)

# Total and percentage of land area of RI within flood zones
ri_fhza_2840_land %>% 
  as.data.frame() %>% 
  group_by(Interval) %>% 
  summarize(SqKm = round(as.numeric(sum(Area)/10^6),1), 
            SqMi = round(as.numeric(SqKm/2.59),1),
            PctArea = paste0(as.character(round(as.numeric(sum(Area)/ri_area*100),1)),"%")) %>% 
  kableExtra::kable(longtable = T, booktabs = T,
                    format.args = list(big.mark = ','),
                    caption = "Rhode Island Land Area within Flood Zones",
                    digits = 0, align = "r") %>% 
  kableExtra::kable_styling(latex_options = c("repeat_header"))
```

These percentages closely approximate the proportion of the total population in the state that are within these flood zones. However, exposure varies significantly by population subgroup. Table \@ref(tab:tabNFHZApop) below shows the total and percentages of the general population and various subgroups within flood zones. Aside from the total population, the largest subpopulation within a flood zone are low income persons, followed by individuals over age 64 and under 18. The latter groups are populations of concern who may have limited mobility in the event of an evacuation. 

```{r tabNFHZApop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Rhode Island Populations within Flood Zones."}
# Show table of pops within flood zones
ri_FloodPops_df %>% 
  mutate(PctFlood = paste0(as.character(round(PctFlood,1)),"%")) %>% 
  arrange(-FloodPop) %>% 
  kableExtra::kable(longtable = T, booktabs = T,
                    format.args = list(big.mark = ','),
                    caption = "Rhode Island Populations within Flood Zones",
                    digits = 0, align = "r") %>% 
  kableExtra::kable_styling(latex_options = c("repeat_header"))
```

Some groups are disproportionately exposed to these flood hazards. While approximately 10% of the general population are within flood zones, individuals with serious disability, the elderly (persons over age 64), and households without access to a car exceed the population average within flood zones. These populations of concern are overrepresented in terms of flood exposure (see Figure \@ref(fig:chartNFHZApop) below).

```{r chartNFHZApop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Rhode Island Populations within Flood Zones."}
# Create lollipop plot of pops within flood zones
ri_FloodPops_df %>% 
  ggplot(aes(x = reorder(Group,-PctFlood), 
             y = PctFlood)) +
  geom_segment(aes(x = reorder(Group,-PctFlood), 
                   xend = reorder(Group,-PctFlood),
                   y = ri_FloodPops_df[1,4], yend = PctFlood), 
               color = "skyblue") +
  geom_point(color = "blue", size = 4, alpha = 0.8) +
  coord_flip() + xlab("") + ylab("") + 
  ggtitle("Rhode Island Populations within Flood Zones") + theme_light() +
  theme(panel.grid.major.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.y = element_blank()) +
  geom_text(aes(x = Group, y = PctFlood + 0.2 * sign(PctFlood), 
                label = paste0(round(PctFlood,1),"%")), 
            hjust = 0.1, vjust = -0.5, size = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = ri_FloodPops_df[1,4], linetype = "dashed") +
  geom_text(aes(x = "Disabled", y = 9.1, label = "Below state avg"),
            color = "gray48") +
  geom_segment(aes(x = "No Car HH", xend = "No Car HH", y = 10, yend = 8.2), 
               arrow = arrow(length = unit(0.3,"cm"))) +
  geom_text(aes(x = "Low Income", y = 11.3, label = "Above state avg"),
            color = "gray48") +
  geom_segment(aes(x = "Under 18", xend = "Under 18", y = 10.4, yend = 12.2), 
               arrow = arrow(length = unit(0.3,"cm")))
```

The map below (Figure \@ref(fig:mapNFHZAPopConcern)) shows those overrepresented populations within the flood zones.

```{r mapNFHZAPopConcern, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Rhode Island Populations within Flood Zones."}
# Create a dot density map of transit-dependent populations in flood zone
# Create random points, with 1 point for every 5 people
Over64_nfhza_pts <- ri_blkgrps_nfhza %>% 
  dplyr::select(NewOver64) %>% 
  filter(NewOver64 >= 5) %>% 
  st_sample(., size = round(.$NewOver64/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Over 64")

Disabled_nfhza_pts <- ri_tracts_nfhza %>% 
  dplyr::select(NewDisabled) %>% 
  filter(NewDisabled >= 5) %>% 
  st_sample(., size = round(.$NewDisabled/5)) %>% 
  st_sf(.) %>% 
  mutate(Group = "Disabled")

NoCarHH_nfhza_pts <- ri_tracts_nfhza %>% 
  dplyr::select(NewNoCar) %>% 
  filter(NewNoCar >= 5) %>% 
  st_sample(., size = round(.$NewNoCar/5)) %>% 
  st_sf(.) %>% 
  mutate(Group = "No Car HH")

# Bring them together
ri_nfhza_vulnerable <- rbind(Over64_nfhza_pts,Disabled_nfhza_pts,
                             NoCarHH_nfhza_pts) %>% 
  slice(sample(1:n())) # randomise order to avoid bias in plotting order

# Map transit-dependent pops and flood zones
tm_layout(bg.color = "#e6f3f7") +
  tm_shape(ri_blkgrp_2840, unit = "mi") + tm_fill(col = "white") +
  tm_shape(ne_states_sf_cb) + tm_fill(col="white") +
  tm_shape(ri_awater_sf) + tm_fill(col = "#e6f3f7") + 
  tm_shape(ri_fhza_2840_land) + 
  tm_fill(col = "Interval", 
          palette = c("gold", "goldenrod3"), 
          labels = c("1% AEP (100-year)", "0.2% AEP (500-year)"), 
          title = "FEMA Flood Zones",
          border.alpha = 0) +
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ri_nfhza_vulnerable) + tm_dots(col = "Group", 
                                          palette = c("green","red","blue"),
                                          legend.show = FALSE) +
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ri_highways) + tm_lines(col = "seashell4", lwd = 2) +
  tm_shape(I95roadSegment) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I95roadSegment2) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I195roadSegment) +
  tm_symbols(shape = I195, border.lwd = NA, size = 0.25) +
  tm_shape(I295roadSegment) + 
  tm_symbols(shape = I295, border.lwd = NA, size = 0.25) +
  tm_shape(ri_towns_sf_pts) + tm_dots(size = 1.5) +
  tm_text("NAME", size = 0.5, col = "black", 
          xmod = 0.7, ymod = 0.2, shadow = TRUE) +
  tm_scale_bar(breaks = c(0, 5, 10), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_add_legend(type = "fill", col = c("green","red","blue"),
                border.col = "white", border.alpha = 0,
                labels = c("Disabled", "No Car HH", "Over 64"),
                title = "Population\nGroup*") +
  tm_layout(title = "Transit Dependent Populations \nwithin Flood Zones", 
            frame = TRUE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"))
```

