---
title: "Evacuation Risks in Connecticut"
author: "Marcos Luna and Neenah Estrella-Luna"
date: "3/20/2020"
output: 
  bookdown::pdf_document2:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: yes        
    includes:  
      in_header: my_header.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\pagebreak
# Analysis of evacuation-related risks in Connecticut

This is an analysis of flood or storm surge-related evacuation risks for populations of concern in Connecticut. In the event of significant inland or coastal flooding, evacuation may be required. For individuals and households with limited mobility, either due to inadequate access to transportation options or because of physical limitations, evacuation presents heightened risk. Evacuation may also prove especially difficult for individuals and households due to limited economic resources, difficulty understanding or accessing information, or low trust in official sources of information. 

## Analysis of Flood Hazard Exposure in Connecticut

```{r NFHZAdata, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Analysis of flooding and hurricane evacuation risks for populations of concern in Connecticut
library(tidyverse)
library(sf)
library(tmap)
library(tmaptools)
library(lwgeom)
library(tigris)
options(tigris_use_cache = TRUE, tigris_class = "sf")

load("DATA/ne_layers.rds")

# Extract state census units and convert to projected local CRS EPSG:2775: NAD83(HARN) / Connecticut. See https://spatialreference.org/ref/epsg/2775/
ct_blkgrp_2775 <- ne_blkgrp_sf %>% 
  filter(STATE == "Connecticut") %>% 
  st_transform(., crs = 2775)

ct_tracts_2775 <- ne_tracts_sf %>% 
  filter(STATE == "Connecticut") %>% 
  st_transform(., crs = 2775)

# Get rid of empty geometries
empty_geo <- st_is_empty(ct_blkgrp_2775)
ct_blkgrp_2775 <- ct_blkgrp_2775[!empty_geo,]
empty_geo <- st_is_empty(ct_tracts_2775)
ct_tracts_2775 <- ct_tracts_2775[!empty_geo,]

# Extract state boundary
ct_state_sf <- ne_states_sf_cb %>% 
  filter(NAME == "Connecticut") %>% 
  st_transform(., crs = 2775)

#### Processing in ArcGIS #####
# Write out block groups for processing in ArcGIS
# ct_blkgrp_2775 %>%
#   dplyr::select(GEOID) %>%
#   st_write(., "DATA/FEMA/CT/ct_blkgrp_2775.shp", delete_layer = TRUE)
# 
# # repeat for tracts
# ct_tracts_2775 %>% 
#   dplyr::select(GEOID) %>%
#   st_write(., "DATA/FEMA/CT/ct_tracts_2775.shp", delete_layer = TRUE)
# 
# # repeat for state boundary
# ct_state_sf %>% 
#   st_write(., "DATA/FEMA/CT/ct_state_2775.shp", delete_layer = TRUE)

# Use dasymetric mapping to calculate populations within flood zones. Approach follows method used by Qiang (2019) to eliminate unpopulated areas of census polygons and then reallocate populations to developed areas as identified in National Land Cover Dataset (NLCD). 
# Perform NLCD raster-to-vector conversion, vector erase/difference, and vector intersections in ArcMap because it takes too long in R. 
# In ArcMap:
# Convert NLCD raster to shapefile. Clip to state.
# Isolate undeveloped areas (gridcode NOT 22 - 24). 
# Erase areas of ct_blkgrp_2775 and ct_tracts_2775 that overlap with undeveloped areas in NLCD shapefiles. Compute OldArea of erased polygons in sqm to identify area of developed polygons remaining. 
# Intersect erased ct_blkgrps and erased ct_tracts with NFHZA and Hurricane evacuation zones. Read back into R.

##### Return to working in R ######
# read in processed ct_blkgrps and ct_tracts
# st_layers(dsn = "DATA/FEMA/CT")

ct_blkgrps_nfhza <- st_read(dsn = "DATA/FEMA/CT", 
                            layer = "ct_blkgrps_nfhza") %>% 
  left_join(., as.data.frame(ct_blkgrp_2775), by = "GEOID") %>% 
  st_transform(., crs = 2775) %>% 
  mutate(NewArea = st_area(.)) %>%
  st_make_valid()

ct_tracts_nfhza <- st_read(dsn = "DATA/FEMA/CT", 
                           layer = "ct_tracts_nfhza") %>% 
  left_join(., as.data.frame(ct_tracts_2775), by = "GEOID") %>% 
  st_transform(., crs = 2775) %>% 
  mutate(NewArea = st_area(.)) %>%
  st_make_valid()

# Apportion populations based on geographic proportion of intersect
ct_blkgrps_nfhza <- ct_blkgrps_nfhza %>%
  mutate(CT_LOWINC = if_else(CT_INCOME == "I",totalpopE,0)) %>%
  mutate(CT_LOWINC = replace_na(CT_LOWINC,0)) %>%
  mutate(Proportion = as.numeric(NewArea/OldArea),
         NewPop = totalpopE*Proportion,
         NewMinority = minorityE*Proportion,
         NewUnder5 = under5E*Proportion,
         NewOver64 = over64E*Proportion,
         NewUnder18 = under18E*Proportion,
         NewEng_limit = eng_limitE*Proportion,
         NewPov = num2povE*Proportion,
         NewLths = lthsE*Proportion,
         NewCT_LOWINC = CT_LOWINC*Proportion)

ct_tracts_nfhza <- ct_tracts_nfhza %>%
  mutate(Proportion = as.numeric(NewArea/OldArea),
         NewDisabled = disabledOver18E*Proportion,
         NewNoCar = HHnoCarE*Proportion)

# Compute total block group populations within flood zones
ct_flood_blkgrp_df <- ct_blkgrps_nfhza %>%
  as.data.frame() %>%
  summarize(`Total Pop` = as.integer(sum(NewPop)),
            Minority = as.integer(sum(NewMinority)),
            `Under 5` = as.integer(sum(NewUnder5)),
            `Over 64` = as.integer(sum(NewOver64)),
            `Under 18` = as.integer(sum(NewUnder18)),
            `Limited English HH` = as.integer(sum(NewEng_limit)),
            `Low Income` = as.integer(sum(NewPov)),
            `No HS Dip` = as.integer(sum(NewLths)),
            `CT Low Income` = as.integer(sum(NewCT_LOWINC))) %>%
  gather(key = Group, value = FloodPop)

ct_flood_tracts_df <- ct_tracts_nfhza %>%
  as.data.frame() %>%
  summarize(`Disabled` = as.integer(sum(NewDisabled)),
            `No Car HH` = as.integer(sum(NewNoCar))) %>%
  gather(key = Group, value = FloodPop)

# Compute total tract populations within the state for same groups
ct_tract_flood_pops_df <- ct_tracts_2775 %>%
  as.data.frame() %>%
  summarize(`Disabled` = sum(disabledOver18E),
            `No Car HH` = sum(HHnoCarE)) %>%
  gather(key = Group, value = CTPop) %>%
  left_join(.,ct_flood_tracts_df, by = "Group")

# Compute populations for state,and join with flood pops
ct_FloodPops_df <- ct_blkgrp_2775 %>%
  as.data.frame() %>%
  mutate(CT_LOWINC = if_else(CT_INCOME == "I",totalpopE,0)) %>%
  mutate(CT_LOWINC = replace_na(CT_LOWINC,0)) %>%
  summarize(`Total Pop` = sum(totalpopE),
            Minority = sum(minorityE),
            `Under 5` = sum(under5E),
            `Over 64` = sum(over64E),
            `Under 18` = sum(under18E),
            `Limited English HH` = sum(eng_limitE),
            `Low Income` = sum(num2povE),
            `No HS Dip` = sum(lthsE),
            `CT Low Income` = sum(CT_LOWINC, na.rm = TRUE)) %>%
  gather(key = Group, value = CTPop) %>%
  left_join(., ct_flood_blkgrp_df, by = "Group") %>%
  rbind(.,ct_tract_flood_pops_df) %>%
  mutate(PctFlood = FloodPop/CTPop*100)
```

As a humid, coastal state, a significant portion of Connecticut's land area and population are exposed to the risk of flooding from overbanking of inland water bodies (e.g., ponds and rivers) or from coastal storm surge and sea level rise. 

The analysis of flood exposure presented here is based on the Federal Emergency Management Agency's (FEMA) National Flood Hazard Layer (NFHL), a digital version of FEMA's most recent flood maps.  FEMA identifies areas subject to varying levels of flood risk through its Flood Insurance Rate Maps (FIRMs), which are produced for most parts of the country to identify areas subject to flooding. FEMA's FIRMs are a national standard used by all federal agencies for the purposes of requiring and rating the purchase of flood insurance and regulating new development.  

A significant proportion of Connecticut's population lives near the coast or near inland water bodies. As a result a significant proportion of the population lives near or within known flood zones. Figure \@ref(fig:mapNFHZAPop) below shows the population distribution and areas subject to floods with an Annual Exceedance Probability (AEP) of 1% (also known as a '100-year' flood) and areas subject to 0.2% AEP (also known as a '500-year' flood). Areas within the 1% AEP flood zone are designated by FEMA as Special Flood Hazard Areas, and development within those zones must be covered by flood insurance. Areas within the 0.2% AEP are not currently regulated, but these areas are nevertheless subject to flood risk under more extreme, albeit less frequent, flooding circumstances. Coastal areas within the 0.2% AEP zone are a reasonable proxy measure for the risks of sea level rise. Note that portions of northwest and northeast Connecticut are not currently mapped for flooding risk by FEMA. 

```{r mapNFHZAPop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="FEMA flood zones and population distribution across Connecticut at Census Block Group level."}
# Create a dot density map of total populations and overlay on flood zones
# Create point layer of major cities for context
# Note cb=FALSE is necessary for extracting centroids from town polygons. Otherwise, if cb=TRUE, cannot extract centroids from multipolygon features.
# Read in NFHL for CT. Data comes from FEMA. 
# List available layers in geodatabase
# st_layers("DATA/FEMA/CT/NFHL_09_20200116/NFHL_09_20200116.gdb")
# Read in flood hazard areas
ct_nfhza_2775 <- st_read(dsn = "DATA/FEMA/CT/NFHL_09_20200116/NFHL_09_20200116.gdb", layer = "S_FLD_HAZ_AR") %>% 
  filter(!FLD_ZONE %in% c("OPEN WATER","AREA NOT INCLUDED") & 
           !ZONE_SUBTY %in% c("AREA OF MINIMAL FLOOD HAZARD",
                              "AREA WITH REDUCED FLOOD RISK DUE TO LEVEE")) %>%
  mutate(FLD_ZONE = as.character(FLD_ZONE)) %>% # omit unused factor levels
  st_transform(., crs = 2775) %>% 
  st_make_valid() %>% 
  group_by(FLD_ZONE) %>% # aggregate flood zone polygons
  summarize(count = n()) %>% 
  mutate(Area = st_area(.),
         Interval = case_when(
           FLD_ZONE  == "A" ~ "100-year",
           FLD_ZONE  == "AE" ~ "100-year",
           FLD_ZONE  == "AH" ~ "100-year",
           FLD_ZONE  == "AO" ~ "100-year",
           FLD_ZONE  == "VE" ~ "100-year",
           FLD_ZONE == "X" ~ "500-year"))

# get rid of empty geometries
empty_geo <- st_is_empty(ct_nfhza_2775)
ct_nfhza_2775 <- ct_nfhza_2775[!empty_geo,]


# download Census TIGERLine hydrography for CT
## First, extract list of county names to use with tigris::water
ct_counties <- counties("CT") %>% 
  pull(NAME)
# Next, download water features for each county and rbind to one layer
ct_awater_sf <- rbind_tigris(
  lapply(
    ct_counties, function(x) area_water(state = "CT", county = x)
  )
) %>% 
  st_union() %>% 
  st_as_sf() %>% 
  st_transform(., crs = 2775)

# crop flood zones to land areas only
ct_nfhza_2775_land <- ct_nfhza_2775 %>% 
  crop_shape(., ct_state_sf, polygon = TRUE) %>% 
  st_difference(., ct_awater_sf) %>% 
  mutate(Area = st_area(.)) %>% 
  st_make_valid() # takes about 12.2 mins

# Create a dot density map of total populations and overlay on flood zones
# Create point layer of major cities for context
# Note cb=FALSE is necessary for extracting centroids from town polygons. Otherwise, if cb=TRUE, cannot extract centroids from multipolygon features.
ct_towns_sf_pts <- county_subdivisions(state = "CT", cb = TRUE) %>% 
  filter(NAME %in% c("Hartford", 
                     "Stamford", 
                     "Fairfield", 
                     "Milford", 
                     "New Haven", 
                     "Danbury", 
                     "Hamden", 
                     "Meriden", 
                     "Waterbury", 
                     "New Britain",
                     "New London",
                     "Norwich",
                     "North Canaan",
                     "Old Saybrook")) %>% 
  st_transform(., crs = 2775) %>% 
  st_centroid(of_largest_polygon = TRUE)

# Create road layer for context
ct_highways <- primary_roads() %>% 
  filter(FULLNAME %in% c("I- 84","I- 91","I- 95","I- 395")) %>% 
  tmaptools::crop_shape(., ne_states_sf_cb) %>% 
  st_transform(., crs = 2775)

# Extract highway segments for labeling
I84roadSegment <- ct_highways %>% 
  filter(LINEARID == "1105281313426")

I91roadSegment <- ct_highways %>% 
  filter(LINEARID == "1104470395750")

I95roadSegment <- ct_highways %>% 
  filter(LINEARID == "1105068628699")

I95roadSegment2 <- ct_highways %>% 
  filter(LINEARID == "1105281377007")

I395roadSegment <- ct_highways %>% 
  filter(LINEARID == "1104493001068")

# Create custom icons of highway shields
I95 <- tmap_icons(file = "https://upload.wikimedia.org/wikipedia/commons/thumb/6/61/I-95.svg/200px-I-95.svg.png")
I395 <- tmap_icons("https://upload.wikimedia.org/wikipedia/commons/thumb/1/16/I-395.svg/200px-I-395.svg.png")
I91 <- tmap_icons("https://upload.wikimedia.org/wikipedia/commons/thumb/9/90/I-91.svg/200px-I-91.svg.png")
I84 <- tmap_icons("https://upload.wikimedia.org/wikipedia/commons/thumb/5/51/I-84.svg/200px-I-84.svg.png")

# Bring in outline of New York for western boundary
ny_state_sf <- tigris::states(cb = TRUE) %>% 
  filter(NAME == "New York")

# Create polygons to show where there are no FEMA maps
nofemapolys <- st_read(dsn = "DATA/FEMA/CT/NFHL_09_20200116/NFHL_09_20200116.gdb", layer = "S_FLD_HAZ_AR") %>% 
  group_by() %>% 
  summarize() %>% # aggregate all flood zone polygons into one
  st_transform(., crs = st_crs(ct_state_sf)) %>% 
  st_make_valid() %>% 
  st_difference(ct_state_sf,.) %>% # extract areas with no flood coverage
  st_cast(., "POLYGON") %>% # disaggregate multipolygon feature
  mutate(Area = st_area(.), 
         LABEL = "No FEMA\nFlood Data") %>% # calc area of each polygon
  filter(as.numeric(Area)/10^6 > 1000) # remove polygons < 1000km2

# Create random points, with 1 point for every 500 people
ct_totalpop_pts <- ct_blkgrp_2775 %>% 
  select(totalpopE) %>% 
  filter(totalpopE >= 500) %>% 
  st_sample(., size = round(.$totalpopE/500)) %>% # create 1 random point for every 500 people
  st_sf(.) %>% 
  group_by() %>% 
  summarize() # reduce to one multipoint object

# Create a simplified version of NFHZA polygons to speed up mapping
ct_nfhza_2775_land_simple <- ct_nfhza_2775_land %>% 
  select(Interval) %>% 
  st_cast(., "MULTIPOLYGON") %>% # homogenize type
  st_cast(., "POLYGON") %>% 
  st_simplify(., dTolerance = 100) %>% # reduce vertices
  st_make_valid() %>%
  group_by(Interval) %>% 
  summarize() %>% 
  mutate(Area = st_area(.))

# Map totalpop and flood zones
tm_layout(bg.color = "#e6f3f7") +
  tm_shape(ct_blkgrp_2775, unit = "mi") + tm_fill(col = "white") +
  tm_shape(ne_states_sf_cb) + tm_fill(col="white") +
  tm_shape(ny_state_sf) + tm_fill(col="white") +
  tm_shape(ct_awater_sf) + tm_fill(col = "#e6f3f7") + 
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ny_state_sf) + tm_borders() +
  tm_shape(ct_totalpop_pts) + tm_dots(col = "darkgoldenrod3",
                                      labels = "1 dot = 500 people",
                                      alpha = 0.6) +
  tm_shape(ct_nfhza_2775_land_simple) + 
  tm_fill(col = "Interval", 
          palette = c("dodgerblue", "blueviolet"), 
          labels = c("1% AEP (100-year)", "0.2% AEP (500-year)"), 
          title = "FEMA Flood Zones",
          alpha = 0.6,
          border.alpha = 0) +
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ny_state_sf) + tm_borders() +
  tm_shape(ct_highways) + tm_lines(col = "seashell4", lwd = 2) +
  tm_shape(I95roadSegment) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I95roadSegment2) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I395roadSegment) +
  tm_symbols(shape = I395, border.lwd = NA, size = 0.25) +
  tm_shape(I91roadSegment) + 
  tm_symbols(shape = I91, border.lwd = NA, size = 0.25) +
  tm_shape(I84roadSegment) + 
  tm_symbols(shape = I84, border.lwd = NA, size = 0.25) +
  tm_shape(ct_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", 
          xmod = 0.7, ymod = 0.2, shadow = TRUE) +
  tm_shape(nofemapolys) + tm_fill(col = "gray", alpha = 0.4) +
  tm_text("LABEL", alpha = 0.5) +
  tm_scale_bar(breaks = c(0, 10, 20), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_add_legend(type = "fill", col = "darkgoldenrod3",
                alpha = 0.6,
                border.col = "white", border.alpha = 0,
                labels = "1 dot = 500 people",
                title = "Total Population") +
  tm_layout(title = "Population Distribution \nand Flood Zones", 
            frame = TRUE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"))
```

Approximately 6.7% of Rhode Island's land area falls within FEMA flood zones. The breakdown by type of flood zone is presented in Table \@ref(tab:tabNFHZAarea) below. 

```{r tabNFHZAarea, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Connecticut Land Area within Flood Zones."}
# Total land in CT
ct_area <- as.numeric(ct_state_sf$ALAND)

# Total and percentage of land area of RI within flood zones
ct_nfhza_2775_land %>% 
  as.data.frame() %>% 
  group_by(Interval) %>% 
  summarize(SqKm = round(as.numeric(sum(Area)/10^6),1), 
            SqMi = round(as.numeric(SqKm/2.59),1),
            PctArea = paste0(as.character(round(as.numeric(sum(Area)/ct_area*100),1)),"%")) %>% 
  kableExtra::kable(longtable = T, booktabs = T,
                    format.args = list(big.mark = ','),
                    caption = "Connecticut Land Area within Flood Zones",
                    digits = 0, align = "r") %>% 
  kableExtra::kable_styling(latex_options = c("repeat_header"))
```

These percentages approximate the proportion of the total population in the state that are within these flood zones. However, exposure varies significantly by population subgroup. Table \@ref(tab:tabNFHZApop) below shows the total and percentages of the general population and various subgroups within flood zones. Aside from the total population, the largest subpopulations within a flood zone are minorities, followed by low income persons. The latter groups are populations of concern who may have limited mobility in the event of an evacuation. 

```{r tabNFHZApop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Connecticut Populations within Flood Zones."}
# Show table of pops within flood zones
ct_FloodPops_df %>% 
  mutate(PctFlood = paste0(as.character(round(PctFlood,1)),"%")) %>% 
  arrange(-FloodPop) %>% 
  kableExtra::kable(longtable = T, booktabs = T,
                    format.args = list(big.mark = ','),
                    caption = "Connecticut Populations within Flood Zones",
                    digits = 0, align = "r") %>% 
  kableExtra::kable_styling(latex_options = c("repeat_header"))
```

Some groups are disproportionately exposed to these flood hazards. While approximately 7.2% of the general population are within flood zones, households without access to a car, people in low income households, limited English speaking households, and the elderly (persons over age 64), exceed the population average within flood zones. These populations of concern are overrepresented in terms of flood exposure (see Figure \@ref(fig:chartNFHZApop) below).

```{r chartNFHZApop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Connecticut Populations within Flood Zones."}
# Create lollipop plot of pops within flood zones
ct_FloodPops_df %>% 
  ggplot(aes(x = reorder(Group,PctFlood), 
             y = PctFlood)) +
  geom_segment(aes(x = reorder(Group,PctFlood), 
                   xend = reorder(Group,PctFlood),
                   y = ct_FloodPops_df[1,4], yend = PctFlood), 
               color = "skyblue") +
  geom_point(color = "blue", size = 4, alpha = 0.8) +
  coord_flip() + xlab("") + ylab("") + 
  ggtitle("Connecticut Populations within Flood Zones") + theme_light() +
  theme(panel.grid.major.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.y = element_blank()) +
  geom_text(aes(x = Group, y = PctFlood + 0.2 * sign(PctFlood), 
                label = paste0(round(PctFlood,1),"%")), 
            hjust = 1.8, vjust = -1, size = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = ct_FloodPops_df[1,4], linetype = "dashed") +
  geom_text(aes(x = "No HS Dip", y = 7.65, label = "Above state avg"),
            color = "gray48") +
  geom_segment(aes(x = "Total Pop", xend = "Total Pop", y = 7.2, yend = 7.6), 
               arrow = arrow(length = unit(0.3,"cm")))
```

The maps below (Figure \@ref(fig:mapNFHZAPopConcern)) show those overrepresented populations within the flood zones.

```{r mapNFHZAPopConcern, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Rhode Island Populations within Flood Zones."}
# Create a dot density map of transit-dependent populations in flood zone
# Create random points, with 1 point for every 5 people
NoCarHH_nfhza_pts <- ct_tracts_nfhza %>% 
  dplyr::select(NewNoCar) %>% 
  filter(NewNoCar >= 5) %>% 
  st_sample(., size = round(.$NewNoCar/5)) %>% 
  st_sf(.) %>% 
  mutate(Group = "No Car HH")

CTLowInc_nfhza_pts <- ct_blkgrps_nfhza %>% 
  select(NewCT_LOWINC) %>% 
  filter(NewCT_LOWINC >= 5) %>% 
  st_sample(., size = round(.$NewCT_LOWINC/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "CT Low Income")

LimitEngHH_nfhza_pts <- ct_blkgrps_nfhza %>% 
  select(NewEng_limit) %>% 
  filter(NewEng_limit >= 5) %>% 
  st_sample(., size = round(.$NewEng_limit/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Limited English HH")

Disabled_nfhza_pts <- ct_tracts_nfhza %>% 
  dplyr::select(NewDisabled) %>% 
  filter(NewDisabled >= 5) %>% 
  st_sample(., size = round(.$NewDisabled/5)) %>% 
  st_sf(.) %>% 
  mutate(Group = "Disabled")

LowInc_nfhza_pts <- ct_blkgrps_nfhza %>% 
  select(NewPov) %>% 
  filter(NewPov >= 5) %>% 
  st_sample(., size = round(.$NewPov/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Low Income")

Minor_nfhza_pts <- ct_blkgrps_nfhza %>% 
  select(NewMinority) %>% 
  filter(NewMinority >= 5) %>% 
  st_sample(., size = round(.$NewMinority/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Minority")

Over64_nfhza_pts <- ct_blkgrps_nfhza %>% 
  dplyr::select(NewOver64) %>% 
  filter(NewOver64 >= 5) %>% 
  st_sample(., size = round(.$NewOver64/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Over 64")

Under5_nfhza_pts <- ct_blkgrps_nfhza %>% 
  select(NewUnder5) %>% 
  filter(NewUnder5 >= 5) %>% 
  st_sample(., size = round(.$NewUnder5/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Under 5")

NoHSdip_nfhza_pts <- ct_blkgrps_nfhza %>% 
  select(NewLths) %>% 
  filter(NewLths >= 5) %>% 
  st_sample(., size = round(.$NewLths/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "No HS Dip")

# Bring them together
ct_nfhza_vulnerable <- rbind(NoCarHH_nfhza_pts, 
                             CTLowInc_nfhza_pts, 
                             LimitEngHH_nfhza_pts,
                             Disabled_nfhza_pts,
                             LowInc_nfhza_pts,
                             Minor_nfhza_pts,
                             Over64_nfhza_pts,
                             Under5_nfhza_pts,
                             NoHSdip_nfhza_pts) %>% 
  # slice(sample(1:n())) %>% # randomise order to avoid bias in plotting order
  group_by(Group) %>% 
  summarize()

# clean up pts layers
rm(list = ls(pattern = "_nfhza_pts"))

# Map over-represented pops and flood zones
tm_layout(bg.color = "#e6f3f7") +
  tm_shape(ct_blkgrp_2775, unit = "mi") + tm_fill(col = "white") +
  tm_shape(ne_states_sf_cb) + tm_fill(col="white") +
  tm_shape(ny_state_sf) + tm_fill(col="white") +
  tm_shape(ct_awater_sf) + tm_fill(col = "#e6f3f7") + 
  tm_shape(ct_nfhza_vulnerable) + 
  tm_dots(col = "darkgoldenrod3", 
          labels = "1 dot = 500 people",
          alpha = 0.6) +
  tm_facets(by = "Group") +
  tm_shape(ct_nfhza_2775_land_simple) + 
  tm_fill(col = "Interval", 
          palette = c("dodgerblue", "blueviolet"), 
          labels = c("1% AEP (100-year)", "0.2% AEP (500-year)"), 
          title = "FEMA Flood Zones",
          alpha = 0.6,
          border.alpha = 0) +
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ny_state_sf) + tm_borders() +
  tm_shape(ct_highways) + tm_lines(col = "seashell4", lwd = 2) +
  tm_shape(I95roadSegment) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I95roadSegment2) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I395roadSegment) +
  tm_symbols(shape = I395, border.lwd = NA, size = 0.25) +
  tm_shape(I91roadSegment) + 
  tm_symbols(shape = I91, border.lwd = NA, size = 0.25) +
  tm_shape(I84roadSegment) + 
  tm_symbols(shape = I84, border.lwd = NA, size = 0.25) +
  tm_shape(ct_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", 
          xmod = 0.7, ymod = 0.2, shadow = TRUE) +
  tm_shape(nofemapolys) + tm_fill(col = "gray", alpha = 0.4) +
  tm_text("LABEL", alpha = 0.5) +
  tm_scale_bar(breaks = c(0, 10, 20), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_add_legend(type = "fill", col = "darkgoldenrod3",
                alpha = 0.6,
                border.col = "white", border.alpha = 0,
                labels = "1 dot = 5 people",
                title = "Population\nGroup") +
  tm_layout(title = "Over-represented\nPopulations\nwithin\nFlood Zones", 
            frame = TRUE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"))
```

\pagebreak
## Analysis of Hurricane Evacuation Risk
```{r HEVACdata, echo=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# read in hurricane evacuation zone layer
ct_hea_sf <- st_read(dsn = "DATA/FEMA/CT/Hurricane Evacuation Zones (2014)", 
                     layer = "geo_export_ca0640ec-4b94-4f5e-b1be-5a02638d3722") %>% 
  mutate(zone = as.character(zone)) %>% 
  filter(zone %in% c("A","B")) %>% 
  st_transform(., crs = 2775) %>% 
  st_make_valid() %>% 
  mutate(Area = st_area(.))

# read in processed ct_blkgrps and ct_tracts
ct_blkgrps_hevac <- st_read(dsn = "DATA/FEMA/CT", 
                            layer = "ct_blkgrps_hevac") %>% 
  left_join(., as.data.frame(ct_blkgrp_2775), by = "GEOID") %>%
  st_transform(., crs = 2775) %>% 
  mutate(NewArea = st_area(.)) %>%
  st_make_valid()

ct_tracts_hevac <- st_read(dsn = "DATA/FEMA/CT", 
                           layer = "ct_tracts_hevac") %>% 
  left_join(., as.data.frame(ct_tracts_2775), by = "GEOID") %>%
  st_transform(., crs = 2775) %>% 
  mutate(NewArea = st_area(.)) %>% 
  st_make_valid()

# Apportion populations based on geographic proportion of intersect
ct_blkgrps_hevac <- ct_blkgrps_hevac %>%
  mutate(CT_LOWINC = if_else(CT_INCOME == "I",totalpopE,0)) %>%
  mutate(CT_LOWINC = replace_na(CT_LOWINC,0)) %>%
  mutate(Proportion = as.numeric(NewArea/OldArea),
         NewPop = totalpopE*Proportion,
         NewMinority = minorityE*Proportion,
         NewUnder5 = under5E*Proportion,
         NewOver64 = over64E*Proportion,
         NewUnder18 = under18E*Proportion,
         NewEng_limit = eng_limitE*Proportion,
         NewPov = num2povE*Proportion,
         NewLths = lthsE*Proportion,
         NewCT_LOWINC = CT_LOWINC*Proportion)

ct_tracts_hevac <- ct_tracts_hevac %>%
  mutate(Proportion = as.numeric(NewArea/OldArea),
         NewDisabled = disabledOver18E*Proportion,
         NewNoCar = HHnoCarE*Proportion)

# Compute total populations within hurricane evac zones
ct_hevac_blkgrp_df <- ct_blkgrps_hevac %>%
  as.data.frame() %>%
  summarize(`Total Pop` = as.integer(sum(NewPop)),
            Minority = as.integer(sum(NewMinority)),
            `Under 5` = as.integer(sum(NewUnder5)),
            `Over 64` = as.integer(sum(NewOver64)),
            `Under 18` = as.integer(sum(NewUnder18)),
            `Limited English HH` = as.integer(sum(NewEng_limit)),
            `Low Income` = as.integer(sum(NewPov)),
            `No HS Dip` = as.integer(sum(NewLths)),
            `CT Low Income` = as.integer(sum(NewCT_LOWINC))) %>%
  gather(key = Group, value = HevacPop)

ct_hevac_tracts_df <- ct_tracts_hevac %>%
  as.data.frame() %>%
  summarize(`Disabled` = as.integer(sum(NewDisabled)),
            `No Car HH` = as.integer(sum(NewNoCar))) %>%
  gather(key = Group, value = HevacPop)

# Compute total tract populations within the state for same groups
ct_tract_hevac_pops_df <- ct_tracts_2775 %>%
  as.data.frame() %>%
  summarize(`Disabled` = sum(disabledOver18E),
            `No Car HH` = sum(HHnoCarE)) %>%
  gather(key = Group, value = CTPop) %>%
  left_join(.,ct_hevac_tracts_df, by = "Group")

# Compute populations for state,  and join with hurricane evac pops
ct_HevacPops_df <- ct_blkgrp_2775 %>%
  as.data.frame() %>%
  mutate(CT_LOWINC = if_else(CT_INCOME == "I",totalpopE,0)) %>%
  mutate(CT_LOWINC = replace_na(CT_LOWINC,0)) %>%
  summarize(`Total Pop` = sum(totalpopE),
            Minority = sum(minorityE),
            `Under 5` = sum(under5E),
            `Over 64` = sum(over64E),
            `Under 18` = sum(under18E),
            `Limited English HH` = sum(eng_limitE),
            `Low Income` = sum(num2povE),
            `No HS Dip` = sum(lthsE),
            `CT Low Income` = sum(CT_LOWINC, na.rm = TRUE)) %>%
  gather(key = Group, value = CTPop) %>%
  left_join(., ct_hevac_blkgrp_df, by = "Group") %>%
  rbind(.,ct_tract_hevac_pops_df) %>%
  mutate(PctHevac = HevacPop/CTPop*100)
```

Connecticut is subject to significant hurricane risk. Since 1900, Connecticut has been struck by hurricanes 5 times, and 3 of those were major hurricanes (i.e., Category 3 or higher). The most recent hurricane to hit Connecticut directly was Hurricane Gloria in 1985, a Category 2 storm. The latter caused over $91 million in damages in Connecticut alone (mostly around Hartford) and left 727,000 people in Connecticut without power. Hurricane Sandy in 2012 did not hit Connecticut direclty, but nevertheless resulted in flooding Connecticut coastal communities. In general, Connecticut is subject to a [hurricane return period of approximately 17 years](https://www.nhc.noaa.gov/climo/#returns), a level of risk comparable to the Gulf Coast of Texas.  

Hurricane intensity is ranked from 1 to 5 on the Saffir-Simpson scale, which is based on sustained wind speeds. However the most destructive and deadly aspects of hurricanes are due to storm surge - ocean water pushed inland by the high winds. This coastal storm surge may be combined with intense rainfall which can exacerbate flooding. As a result, coastal evacuation may be necessary as a hurricane approaches.

The analysis of hurricane evacuation risk presented here is based on Hurricane Evacuation Maps produced by the U.S. Army Corps of Engineers for New England. The maps show Hurricane Evacuation Zones that are recommended to be evacuated  during potential worst-case hurricane storm surge inundation. These evacuation zones are based on [storm surge modeling by the National Weather Service's National Hurricane Center](https://www.nhc.noaa.gov/nationalsurge/) using the hydrodynamic Sea, Lake, and Overland Surges from Hurricanes (SLOSH) model to simulate storm surge from tropical cyclones. These data products are provided to federal, state, and local government agencies to assist in planning, risk assessment studies, and operational decision-making.

A significant proportion of Connecticut's population lives near the coast and is therefore within these hurricane evacuation zones. Figure \@ref(fig:mapHEVACPop) below shows the population distribution and areas within Hurricane Evacuation Zones. Evacuation Zone “A” is recommended to be evacuated prior to an expected category 1 or 2 hurricane.  Evacuation Zone “B” is recommended to be evacuated prior to an expected category 3 or 4 hurricane.

```{r mapHEVACPop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="FEMA flood zones and population distribution across Connecticut at Census Block Group level."}
# Create a simplified version of NFHZA polygons to speed up mapping
ct_hea_sf_simple <- ct_hea_sf %>% 
  select(zone) %>% 
  st_cast(., "MULTIPOLYGON") %>% # homogenize type
  st_cast(., "POLYGON") %>% 
  st_simplify(., dTolerance = 10) %>% # reduce vertices
  st_make_valid() %>%
  group_by(zone) %>% 
  summarize() %>% 
  mutate(Area = st_area(.))

# Map totalpop and hurricane evacuation zones
tm_layout(bg.color = "#e6f3f7") +
  tm_shape(ct_blkgrp_2775, unit = "mi") + tm_fill(col = "white") +
  tm_shape(ne_states_sf_cb) + tm_fill(col="white") +
  tm_shape(ny_state_sf) + tm_fill(col="white") +
  tm_shape(ct_awater_sf) + tm_fill(col = "#e6f3f7") + 
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ny_state_sf) + tm_borders() +
  tm_shape(ct_totalpop_pts) + 
  tm_dots(col = "darkgoldenrod3",
          labels = "1 dot = 500 people",
          alpha = 0.6) +
  tm_shape(ct_hea_sf_simple) + 
  tm_fill(col = "zone", 
          palette = c("deeppink", "purple3"), 
          labels = c("A: Category 1 - 2", 
                     "B: Category 3 - 4"), 
          title = "Evacuation Zone and\nHurricane Category",
          alpha = 0.6,
          border.alpha = 0) +
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ny_state_sf) + tm_borders() +
  tm_shape(ct_highways) + tm_lines(col = "seashell4", lwd = 2) +
  tm_shape(I95roadSegment) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I95roadSegment2) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I395roadSegment) +
  tm_symbols(shape = I395, border.lwd = NA, size = 0.25) +
  tm_shape(I91roadSegment) + 
  tm_symbols(shape = I91, border.lwd = NA, size = 0.25) +
  tm_shape(I84roadSegment) + 
  tm_symbols(shape = I84, border.lwd = NA, size = 0.25) +
  tm_shape(ct_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", 
          xmod = 0.7, ymod = 0.2, shadow = TRUE) +
  tm_scale_bar(breaks = c(0, 10, 20), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_add_legend(type = "fill", col = "darkgoldenrod3",
                alpha = 0.6,
                border.col = "white", border.alpha = 0,
                labels = "1 dot = 500 people",
                title = "Total Population") +
  tm_layout(title = "Population Distribution \nand Hurricane\nEvacuation Zones", 
            frame = TRUE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"))
```

Approximately 3.3% of Rhode Island's land area falls within hurricane evacuation zones. The breakdown by hurricane evacuation zone is presented in Table \@ref(tab:tabHEVACarea) below.

```{r tabHEVACarea, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Connecticut Populations within Hurricane Evacuation Zones."}
# Total and percentage of land area of RI within hurricane evacuation zones
ct_hea_sf %>% 
  as.data.frame() %>% 
  group_by(zone) %>% 
  summarize(SqKm = round(as.numeric(sum(Area)/10^6),1), 
            SqMi = round(as.numeric(SqKm/2.59),1),
            PctArea = paste0(as.character(round(as.numeric(sum(Area)/ct_area*100),1)),"%")) %>% 
  kableExtra::kable(longtable = T, booktabs = T,
                    format.args = list(big.mark = ','),
                    caption = "Connecticut Land Area within Hurricane Evacuation Zones",
                    digits = 0, align = "r") %>% 
  kableExtra::kable_styling(latex_options = c("repeat_header"))
```

These percentages belie the proportion of the population in the state that are within these evacuation zones. Moreover, exposure varies significantly by population subgroup. Table \@ref(tab:tabHEVACpop) below shows the total and percentages of the general population and various subgroups within hurricane evacuation zones. Aside from the total population, the largest subpopulation within an evacuation zone are minorities, and low income persons, followed by individuals over age 64 and under 18. The latter groups are populations of concern who may have limited mobility in the event of an evacuation. 
\pagebreak
```{r tabHEVACpop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Connecticut Populations within Hurricane Evacuation Zones."}
# Show table of pops within hurricane evacuation zones
ct_HevacPops_df %>% 
  mutate(PctHevac = paste0(as.character(round(PctHevac,1)),"%")) %>% 
  arrange(-HevacPop) %>% 
  kableExtra::kable(longtable = T, booktabs = T,
                    format.args = list(big.mark = ','),
                    caption = "Connecticut Populations within Hurricane Evacuation Zones",
                    digits = 0, align = "r") %>% 
  kableExtra::kable_styling(latex_options = c("repeat_header"))
```

Some groups are disproportionately exposed to these hurricane evacuation hazards. While approximately 9.5% of the general population are within hurricane evacuation zones, households without access to a car, people in low income households, people in limited English speaking households, the elderly (persons over age 64), adults with less than a high school education, individuals with serious disability, and other popultions of concern exceed the population average within hurricane evacuation zones. These populations of concern are overrepresented in terms of hurricane evacuation exposure (see Figure \@ref(fig:chartHEVACpop) below).

```{r chartHEVACpop, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Connecticut Populations within Hurricane Evacuation Zones."}
# Create lollipop plot of pops within flood zones
ct_HevacPops_df %>% 
  ggplot(aes(x = reorder(Group,PctHevac), 
             y = PctHevac)) +
  geom_segment(aes(x = reorder(Group,PctHevac), 
                   xend = reorder(Group,PctHevac),
                   y = ct_HevacPops_df[1,4], yend = PctHevac), 
               color = "skyblue") +
  geom_point(color = "blue", size = 4, alpha = 0.8) +
  coord_flip() + xlab("") + ylab("") + 
  ggtitle("Connecticut Populations within Hurricane Evacuation Zones") + theme_light() +
  theme(panel.grid.major.y = element_blank(),
        panel.border = element_blank(),
        axis.ticks.y = element_blank()) +
  geom_text(aes(x = Group, y = PctHevac + 0.2 * sign(PctHevac), 
                label = paste0(round(PctHevac,1),"%")), 
            hjust = 1.8, vjust = -1, size = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = ct_HevacPops_df[1,4], linetype = "dashed") +
  geom_text(aes(x = "Over 64", y = 10.5, label = "Above state avg"),
            color = "gray48") +
  geom_segment(aes(x = "Total Pop", xend = "Total Pop", y = 9.7, yend = 10.5), 
               arrow = arrow(length = unit(0.3,"cm"))) +
  expand_limits(y = c(9,14.5))
```

The maps below (Figure \@ref(fig:mapHEVACPopConcern)) show those overrepresented populations within the flood zones.

```{r mapHEVACPopConcern, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Connecticut Populations within Hurricane Evacuation Zones."}
# Create a dot density map of over-represented populations in hurricane evacuation zones
# Create random points, with 1 point for every 5 people
NoCarHH_hevac_pts <- ct_tracts_hevac %>% 
  dplyr::select(NewNoCar) %>% 
  filter(NewNoCar >= 5) %>% 
  st_sample(., size = round(.$NewNoCar/5)) %>% 
  st_sf(.) %>% 
  mutate(Group = "No Car HH")

CTLowInc_hevac_pts <- ct_blkgrps_hevac %>% 
  select(NewCT_LOWINC) %>% 
  filter(NewCT_LOWINC >= 5) %>% 
  st_sample(., size = round(.$NewCT_LOWINC/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "CT Low Income")

LimitEngHH_hevac_pts <- ct_blkgrps_hevac %>% 
  select(NewEng_limit) %>% 
  filter(NewEng_limit >= 5) %>% 
  st_sample(., size = round(.$NewEng_limit/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Limited English HH")

Disabled_hevac_pts <- ct_tracts_hevac %>% 
  dplyr::select(NewDisabled) %>% 
  filter(NewDisabled >= 5) %>% 
  st_sample(., size = round(.$NewDisabled/5)) %>% 
  st_sf(.) %>% 
  mutate(Group = "Disabled")

LowInc_hevac_pts <- ct_blkgrps_hevac %>% 
  select(NewPov) %>% 
  filter(NewPov >= 5) %>% 
  st_sample(., size = round(.$NewPov/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Low Income")

Minor_hevac_pts <- ct_blkgrps_hevac %>% 
  select(NewMinority) %>% 
  filter(NewMinority >= 5) %>% 
  st_sample(., size = round(.$NewMinority/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Minority")

Over64_hevac_pts <- ct_blkgrps_hevac %>% 
  dplyr::select(NewOver64) %>% 
  filter(NewOver64 >= 5) %>% 
  st_sample(., size = round(.$NewOver64/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Over 64")

Under5_hevac_pts <- ct_blkgrps_hevac %>% 
  select(NewUnder5) %>% 
  filter(NewUnder5 >= 5) %>% 
  st_sample(., size = round(.$NewUnder5/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "Under 5")

NoHSdip_hevac_pts <- ct_blkgrps_hevac %>% 
  select(NewLths) %>% 
  filter(NewLths >= 5) %>% 
  st_sample(., size = round(.$NewLths/5)) %>% # create 1 random point for every 5 people
  st_sf(.) %>% 
  mutate(Group = "No HS Dip")

# Bring them together
ct_hevac_vulnerable <- rbind(NoCarHH_hevac_pts, 
                             CTLowInc_hevac_pts, 
                             LimitEngHH_hevac_pts,
                             Disabled_hevac_pts,
                             LowInc_hevac_pts,
                             Minor_hevac_pts,
                             Over64_hevac_pts,
                             Under5_hevac_pts,
                             NoHSdip_hevac_pts) %>% 
  # slice(sample(1:n())) %>% # randomise order to avoid bias in plotting order
  group_by(Group) %>% 
  summarize()

# clean up pts layers
rm(list = ls(pattern = "_hevac_pts"))

# Map over-represented pops and hurricane evacuation zones
tm_layout(bg.color = "#e6f3f7") +
  tm_shape(ct_blkgrp_2775, unit = "mi") + tm_fill(col = "white") +
  tm_shape(ne_states_sf_cb) + tm_fill(col="white") +
  tm_shape(ny_state_sf) + tm_fill(col="white") +
  tm_shape(ct_awater_sf) + tm_fill(col = "#e6f3f7") + 
  tm_shape(ct_hevac_vulnerable) + 
  tm_dots(col = "darkgoldenrod3", 
          labels = "1 dot = 500 people",
          alpha = 0.6) +
  tm_facets(by = "Group") +
  tm_shape(ct_hea_sf_simple) + 
  tm_fill(col = "zone", 
          palette = c("deeppink", "purple3"), 
          labels = c("A: Category 1 - 2", 
                     "B: Category 3 - 4"), 
          title = "Evacuation Zone and\nHurricane Category",
          alpha = 0.6,
          border.alpha = 0) +
  tm_shape(ne_states_sf_cb) + tm_borders() +
  tm_shape(ny_state_sf) + tm_borders() +
  tm_shape(ct_highways) + tm_lines(col = "seashell4", lwd = 2) +
  tm_shape(I95roadSegment) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I95roadSegment2) + 
  tm_symbols(shape = I95, border.lwd = NA, size = 0.25) +
  tm_shape(I395roadSegment) +
  tm_symbols(shape = I395, border.lwd = NA, size = 0.25) +
  tm_shape(I91roadSegment) + 
  tm_symbols(shape = I91, border.lwd = NA, size = 0.25) +
  tm_shape(I84roadSegment) + 
  tm_symbols(shape = I84, border.lwd = NA, size = 0.25) +
  tm_shape(ct_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", 
          xmod = 0.7, ymod = 0.2, shadow = TRUE) +
  tm_scale_bar(breaks = c(0, 10, 20), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_add_legend(type = "fill", col = "darkgoldenrod3",
                alpha = 0.6,
                border.col = "white", border.alpha = 0,
                labels = "1 dot = 5 people",
                title = "Population\nGroup") +
  tm_layout(title = "Over-represented\nPopulations\nwithin Hurricane\nEvacuation Zones", 
            frame = TRUE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"))
```
