---
title: "PM~2.5~"
author: "Marcos Luna and Neenah Estrella-Luna"
date: "12/31/2019"
output: 
  bookdown::word_document2:
  fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of PM~2.5~ in New England

This is an analysis of PM~2.5~ across New England and its relationship to populations of concern. The analysis is based on data from the EPA's EJSCREEN. EJSCREEN data provides PM~2.5~ annual concentrations at the Census Block Group level for the years 2011 tp 2016 (as of December 2019). 

PM~2.5~ levels vary significantly across New England, with highest concentrations in southwest Connecticut, from Stamford to Hartford, and also from Providence, Rhode Island, to the Boston metro region (see Figure \@ref(fig:mapPM25NE) below). 

```{r mapPM25NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of 2016 annual PM~2.5~ concentrations across New England at Census Block Group level."}
library(tidyverse)
library(sf)
library(tmap)
library(tmaptools)
library(maptools)
# library(raster)
library(rgdal)
library(RColorBrewer)
library(sp)
library(CGPfunctions) # for slope graphs
library(ggcorrplot)
# ne_blkgrp_sf_DemoEJ %>% 
#   as.data.frame() %>% 
#   transmute(`PM2.5 2011` = pm_15, `PM2.5 2016` = PM25_19) %>% 
#   summary()

# Load data
load("DATA/ne_layers.rds")

# Map of PM2.5 across New England
tm_shape(ne_blkgrp_sf_DemoEJ, unit = "mi") + 
  tm_fill("PM25_19", style = "quantile", 
          title = expression(paste
                             (PM[2.5]," (", mu, "g/", m^3, ")", sep = "")),
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=2)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Annual PM2.5 \nConcentrations\n2016", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

These PM~2.5~ levels vary significantly by state (see Figure \@ref(fig:boxplotPM25NE) below). Connecticut stands out with the highest median PM~2.5~ concentrations among the New England states, while Vermont has the lowest overall concentrations. The differences in concentrations between the states are all statistically significant.  

```{r boxplotPM25NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Boxplot of 2016 annual PM~2.5~ concentrations by state at Census Block Group level."}
# boxplot of PM2.5 by state for 2016
ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  ggplot(aes(x = STATE, y = PM25_19, fill = STATE)) + 
  geom_boxplot(notch = TRUE) + 
  ggtitle(expression(atop(paste(PM[2.5], " Annual Concentration by New England state,"), "2016"))) +
  theme_minimal() +
  theme(legend.position = "none", axis.text=element_text(size=8)) + 
  xlab(NULL) + 
  ylab(expression(paste(PM[2.5]," (", mu, "g/", m^3, ")", sep = ""))) +
  coord_flip()
# # boxplot of PM2.5 by state for 2011
# ne_blkgrp_sf_DemoEJ %>% 
#   as.data.frame() %>% 
#   ggplot(aes(x = STATE, y = pm_15, fill = STATE)) + 
#   geom_boxplot(notch = TRUE) + 
#   ggtitle(expression(paste(PM[2.5], " Annual Concentration by New England state, 2011", sep = ""))) +
#   theme_minimal() +
#   theme(legend.position = "none") + xlab(NULL) + 
#   ylab(expression(paste(PM[2.5]," (", mu, "g/", m^3, ")", sep = "")))
```


Since 2011, PM~2.5~ levels have declined across the region, on average by `r {round(mean(ne_blkgrp_sf_DemoEJ$PM25_pctChange, na.rm = TRUE),1)}`%. Unsurprisingly, this decline has not been uniform (see Figure \@ref(fig:mapPM25NEchange) below). The greatest declines, up to 43%, have been in the northern half of Vermont and the region between Boston and Portsmouth. 

```{r mapPM25NEchange, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Map of percent change in annual PM~2.5~ concentrations across New England between 2011 and 2016 at Census Block Group level."}
# Map of PM2.5 change across New England
tm_shape(ne_blkgrp_sf_DemoEJ, unit = "mi") + 
  tm_fill("PM25_pctChange", palette = "Greens",
          style = "quantile", 
          title = "% Change",
          legend.hist = TRUE,
          colorNA = NULL,
          textNA = NULL,
          legend.format=list(list(digits=1)),
          legend.is.portrait = TRUE) +
  tm_shape(ne_states_sf_cb) + tm_borders(alpha = 0.4) + 
  tm_text("STUSPS", size = 0.7, remove.overlap = TRUE, col = "gray") +
  tm_shape(ne_towns_sf_pts) + tm_dots() +
  tm_text("NAME", size = 0.5, col = "black", xmod = 0.7, ymod = 0.2) +
  tm_scale_bar(breaks = c(0, 50, 100), text.size = 0.5, 
               position = c(0.6,0.005)) +
  tm_layout(title = "Change in annual PM2.5 \nConcentrations \n2011-2016", 
            frame = FALSE, main.title.size = 0.8,
            legend.outside = TRUE,
            legend.title.size = 0.8,
            legend.outside.position = c("right", "top"),
            legend.hist.width = 0.9)
```

Averaged across the states, the differences in these declines are also apparent. Figure \@ref(fig:slopegraphPM25NE) below compares the average annual PM~2.5~ concentrations by state and for the region between 2011 and 2016. The region and the states all showed significant declines since 2011, although note that concentrations in Connecticut have remained consistently well above the rest of the region. 

```{r slopegraphPM25NE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Slope graph of annual PM~2.5~ concentrations for 2011 and 2016 for New England and states."}
# Slope graph of PM2.5 by state and region between 2011 and 2016
# Create df of 2019 actual values
PM19wSTAvgs_actual <- ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(PM25Mean = mean(PM25_19, na.rm = TRUE),
            PM25wMean = weighted.mean(x = PM25_19,
                                      w = totalpopE, na.rm = TRUE)) %>% 
  mutate(Year = 2016)
# Create df of 2015 values
PM15wSTAvgs_actual <- ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  group_by(STATE) %>% 
  summarize(PM25Mean = mean(pm_15, na.rm = TRUE),
            PM25wMean = weighted.mean(x = pm_15,
                                      w = pop_15, na.rm = TRUE)) %>% 
  mutate(Year = 2011)
# Create regional average benchmark
pm15NEavg <- ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  summarize(PM25Mean = mean(pm_15, na.rm = TRUE),
            PM25wMean = weighted.mean(x = pm_15,
                                      w = pop_15, na.rm = TRUE)) %>% 
  transmute(STATE = "NEW ENGLAND", 
            PM25Mean = PM25Mean, 
            PM25wMean = PM25wMean,
            Year = 2011)
pm19NEavg <- ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  summarize(PM25Mean = mean(PM25_19, na.rm = TRUE),
            PM25wMean = weighted.mean(x = PM25_19,
                                      w = totalpopE, na.rm = TRUE)) %>% 
  transmute(STATE = "NEW ENGLAND", 
            PM25Mean = PM25Mean, 
            PM25wMean = PM25wMean,
            Year = 2016)
#rbind tables
PM11_16wSTAvg_actual <- bind_rows(PM15wSTAvgs_actual,
                                PM19wSTAvgs_actual,
                                pm15NEavg,pm19NEavg) %>% 
  mutate(Year = as.factor(Year),
         PM25Mean = round(PM25wMean,2),
         PM25wMean = round(PM25wMean,2))
# make slope graph
newggslopegraph(dataframe = PM11_16wSTAvg_actual, 
                Times = Year, 
                Measurement = PM25Mean, 
                Grouping = STATE,
                Title = expression(paste("Annual Average ", PM[2.5], " Concentrations")),
                SubTitle = expression(paste
                                      (PM[2.5]," (", mu, "g/", m^3, ")", sep = "")),
                Caption = NULL,
                LineColor = c("NEW ENGLAND" = "#000000",
                              "Connecticut" = "#E69F00",
                              "Massachusetts" = "#56B4E9",
                              "Rhode Island" = "#009E73",
                              "New Hampshire" = "#F0E442",
                              "Vermont" = "#0072B2",
                              "Maine" = "#D55E00"),
                LineThickness = 1,
                YTextSize = 3.5,
                DataTextSize = 3,
                WiderLabels = TRUE)
```


## PM~2.5~ in New England and Populations of Concern

In addition to variations in the general geography of PM~2.5~ concentrations, exposure to these pollutants also varies demographically. Figure \@ref(fig:barplotPM25PopAvgNE) below shows population-weighted exposures for populations of concern relative to average PM~2.5~ concentrations for the region. For example, minorities in New England are exposed to PM~2.5~ concentrations that are more than 5% above concentrations for the region as a whole. Similarly, linguistically isolated households are exposed to concentrations 4.9% above the regional average. By contrast, persons over age 64 are, on average, exposed to concentrations of PM~2.5~ at or below the regional average. 

```{r barplotPM25PopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Graph of population-weighted average exposures to PM~2.5~ for populations of concern in New England relative to the regional average."}
# Pop Weighted avg of PM2.5 for all Groups in New England relative to NE average
ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(totalpopE,
                nhwhitepopE,
                minorityE,
                nhblackpopE,
                nhamerindpopE,
                nhasianpopE,
                nhnativhpopE,
                nhotherpopE,
                nh2morepopE,
                hisppopE,
                povknownE,
                num2povE, 
                eng_hhE,
                eng_limitE,
                age25upE,
                lthsE, 
                allAgesE, 
                under5E, 
                over65E, 
                PM25_19) %>% 
  gather(key = Group, value = Pop, totalpopE:over65E) %>% 
  group_by(Group) %>% 
  summarize(PM25wMean = weighted.mean(x = PM25_19, w = Pop, na.rm = TRUE),
            PM25Mean = mean(PM25_19, na.rm = TRUE)) %>% 
  spread(key = Group, value = PM25wMean) %>% 
  transmute(Minority = (minorityE/PM25Mean - 1)*100,
         #Minority_NHW = (minorityE/nhwhitepopE - 1)*100,
         `Lang Isol` = (eng_limitE/PM25Mean - 1)*100,
         Poverty = (num2povE/PM25Mean - 1)*100,
         `No HS` = (lthsE/PM25Mean - 1)*100,
         `Under 5` = (under5E/PM25Mean - 1)*100,
         `Over 65` = (over65E/PM25Mean - 1)*100) %>%
  gather(key = Group, value = Pct) %>% 
  ggplot(aes(x = reorder(Group, -Pct), y = Pct, fill = Group)) + 
  geom_bar(stat = "identity", position = "identity") +
  theme_minimal() +
  labs(x = "", y = "", title = expression(atop(paste("Population-Weighted ", PM[2.5], " Exposure"), "(relative to New England average)"))) + 
  theme(legend.position = 'none') +
  geom_text(aes(x = Group, y = Pct + 0.2 * sign(Pct), 
                label = paste0(round(Pct,2),"%")), 
            hjust = 0.5, size = 3,
            color=rgb(100,100,100, maxColorValue=255)) +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  geom_hline(yintercept = 0)
```

Like the region as a whole, these populations have also experienced a decline in exposure since 2011. The comparison between exposure for these groups since 2011 is displayed below in Figure \@ref(fig:slopegraphPM25PopAvgNE). Note however that except for persons over age 64 and low income persons, all other populations of concern continue to experience exposures greater than the regional average, with minorities leading on this measure. 

```{r slopegraphPM25PopAvgNE, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Slope graph of annual PM~2.5~ concentrations by population of concern for 2011 and 2016."}
# Pop Weighted avg of PM2.5 EXPOSURE CHANGE 2011 - 2016 for all Groups in New England relative to NE average. 
# Create df of 2019 actual values
PM19wAvgs_actual <- ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(minorityE,
                num2povE,
                eng_limitE,
                lthsE,
                under5E,
                over65E,
                PM25_19) %>% 
  gather(key = Group, value = Pop, minorityE:over65E) %>% 
  group_by(Group) %>% 
  summarize(PM25wMean = weighted.mean(x = PM25_19, 
                                      w = Pop, na.rm = TRUE)) %>% 
  mutate(Group = case_when(
    Group == "minorityE" ~ "Minority",
    Group == "eng_limitE" ~ "Lang Isol",
    Group == "num2povE" ~ "Low Income",
    Group == "lthsE" ~ "No HS Dip",
    Group == "under5E" ~ "Under 5",
    Group == "over65E" ~ "Over 64")) %>% 
  mutate(Year = 2016)
# Create df of 2015 values
PM15wAvgs_actual <- ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  # filter(STATE == "Massachusetts") %>%
  dplyr::select(mins_15,
                lowinc_15,
                lingiso_15,
                lths_15,
                under5_15,
                over64_15,
                pm_15) %>% 
  gather(key = Group, value = Pop, mins_15:over64_15) %>% 
  group_by(Group) %>% 
  summarize(PM25wMean = weighted.mean(x = pm_15, 
                                      w = Pop, na.rm = TRUE)) %>%  
  mutate(Group = case_when(
    Group == "mins_15" ~ "Minority",
    Group == "lingiso_15" ~ "Lang Isol",
    Group == "lowinc_15" ~ "Low Income",
    Group == "lths_15" ~ "No HS Dip",
    Group == "under5_15" ~ "Under 5",
    Group == "over64_15" ~ "Over 64")) %>% 
  mutate(Year = 2011)
# Create regional average benchmark
pm15NEavg <- mean(ne_blkgrp_sf_DemoEJ$pm_15, na.rm = TRUE)
pm19NEavg <- mean(ne_blkgrp_sf_DemoEJ$PM25_19, na.rm = TRUE)
pmRegionalAvg_actual <- data.frame(
  Group = c("REGIONAL AVG", "REGIONAL AVG"),
  PM25wMean = c(pm15NEavg,pm19NEavg),
  Year = c(2011,2016)
)
#rbind tables
PM11_16wAvg_actual <- bind_rows(PM15wAvgs_actual,
                                PM19wAvgs_actual,
                                pmRegionalAvg_actual) %>% 
  mutate(Year = as.factor(Year),
         PM25wMean = round(PM25wMean,1))
# make slope graph
newggslopegraph(dataframe = PM11_16wAvg_actual, 
                Times = Year, 
                Measurement = PM25wMean, 
                Grouping = Group,
                Title = "Population-Weighted PM2.5 Exposure",
                SubTitle = expression(paste
                                      (PM[2.5]," (", mu, "g/", m^3, ")", sep = "")),
                Caption = NULL,
                LineColor = c("REGIONAL AVG" = "#000000",
                              "Lang Isol" = "#E69F00",
                              "Minority" = "#56B4E9",
                              "No HS" = "#009E73",
                              "Poverty" = "#F0E442",
                              "Under 5" = "#0072B2",
                              "Over 64" = "#D55E00"),
                YTextSize = 3.5,
                DataTextSize = 3,
                WiderLabels = TRUE)
```

Indeed, there is a moderate but statistically significant positive relationship between the proportion of minority residents or language-isolated households and the concentration of PM~2.5~ (see Figure \@ref(fig:cormatrixPM25PopAvg) below.).

```{r cormatrixPM25PopAvg, echo=FALSE, message=FALSE, warning=FALSE, fig.align = "center", fig.cap="Spearman's correlation matrix of annual PM~2.5~ concentrations and the proportions of populations of concern by Census Block Group."}
# Create a correlation matrix of PM25 and populations of concern
corrPM25 <- ne_blkgrp_sf_DemoEJ %>% 
  as.data.frame() %>% 
  dplyr::transmute(PM25 = PM25_19, 
                Minority = minority_pctE, 
                `Low Income` = pct2povE, 
                `Lang Isol` = eng_limit_pctE, 
                `No HS Dip` = pct_lthsE, 
                `Under 5` = pct_under5E, 
                `Over 64` = pct_over65E) %>% 
  drop_na() %>% 
  cor(method = "spearman")
# corrPM25
ggcorrplot(corrPM25, hc.order = TRUE, type = "lower", lab = TRUE, 
           title = "PM2.5 Correlation Matrix for New England", 
           legend.title = "Spearman's \nCorrelation\nCoefficient")
```

